---
title: "STAT 479 HW1"
output: rmarkdown::pdf_document
header-includes:
    - \usepackage {hyperref}
    - \hypersetup {colorlinks = true, linkcolor = red, urlcolor = red}
---

```{r, echo = FALSE}
library("knitr")
opts_chunk$set(cache = TRUE, message = FALSE, warning = FALSE)
```

## Instructions

1. Submit your solutions to the exercises below before **February 5 at 11:59pm
CST**.
2. Prepare your solutions as Rmarkdown documents. For problems in vega-lite,
copy your code into the `.Rmd`, making sure to enclose them within `js` blocks,

`````
````js
// your code here
````
`````

so that the syntax is properly highlighted, e.g.,

````js
// your code here
````

3. For vega-lite problems, include a screenshot of your result.
4. We give example figures below to guide your work. However, view these only as
suggestions -- we will keep an eye out for improvements over our plots.
5. Include two files in your submission, (a) a pdf of the compiled `.Rmd` file
and (b) the original `.Rmd` file.

## Rubric

2 problems below will be graded according to,

* Correctness: For plots, the displays meet the required specifications. For
conceptual questions, all parts are accurately addressed.
* Attention to detail: Writing is clear and designs are elegant. For example, no
superfluous marks are included, all axes are labeled, text is neither too small
nor too large.
* Code quality (if applicable): Code is concise but readable, properly formatted
and commented.

The remaining problems will be graded for completeness.

## Problems

### (1) Ikea Furniture

The dataset below shows prices of pieces of Ikea furniture. We will compare
prices of different furniture categories and label the (relatively few) articles
which cannot be bought online.

```{r}
library("readr")
ikea <- read_csv("https://uwmadison.box.com/shared/static/iat31h1wjg7abhd2889cput7k264bdzd.csv")
```
a. Make a plot that shows the relationship between the `category` of furniture
and `price` (on a log-scale). Show each `item_id` as a point -- do not aggregate
to boxplots or ridgelines -- but make sure to jitter and adjust the size the
points to reduce the amount of overlap. _Hint: use the `geom_jitter` layer._

b. Modify the plot in (a) so that categories are sorted from those with highest
to lowest average prices.

c. Color points according to whether they can be purchased online. If they
cannot be purchased online, add a text label giving the name of that item of
furniture. An example result is given by Figure 1.

### (2) Penguins

The data below measures properties of various antarctic penguins.

```{r}
penguins <- read_csv("https://uwmadison.box.com/shared/static/ijh7iipc9ect1jf0z8qa2n3j7dgem1gh.csv")
```
Using either vega-lite or ggplot2, create a single plot that makes it easy to
answer both of these questions,

i) How is bill length related to bill depth within and across species? 
ii) On which islands are which species found?

(Notice that the answer to part (i) is an example of Simpson's paradox!)

### (3) 2012 London Olympics

This exercise is similar to the Ikea furniture one, except that it will be
interactive. The data at this
[link](https://uwmadison.box.com/s/rzw8h2x6dp5693gdbpgxaf2koqijo12l) describes
all participants in the London 2012 Olympics. From an observable notebook, the
following code can be used to derive a new variable with a jittered Age
variable, which will be useful in part (a).

````js
import { vl } from "@vega/vega-lite-api"
import { aq, op } from "@uwdata/arquero"
data_raw = aq.fromCSV(await FileAttachment("All London 2012 athletes - ALL ATHLETES.csv").text())
data = data_raw.derive({Age_: d => d.Age + 0.25 * Math.random() })
````

a. Create a layered display that shows (i) the ages of athletes across sports
and (ii) the average age within each sport. Use different marks for participants
and for averages. To avoid overplotting, use the jittered `Age_` variable
defined in the code block above.

b. Sort the sports from lowest to highest average age. Add a tooltip so that
hovering over an athlete shows their name. Your results should look something
like the display in Figure \ref{fig:3}.

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{/Users/kris/Desktop/olympics.png}
  \caption{Example vega-lite result for Problem (3). Hovering over a tick mark
  shows the name of the athlete.}
  \label{fig:3}
\end{figure}

### (4) Traffic

In lecture, we looked at the `geom_density_ridges` function. In this exercise,
we will instead use `geom_ridgeline`, which is useful whenever the heights of the
ridges have been computed in advance. We will use the traffic data read in
below.

```{r}
traffic <- read_csv("https://uwmadison.box.com/shared/static/x0mp3rhhic78vufsxtgrwencchmghbdf.csv")
```

Each row is a timepoint of traffic within a city in Germany. Using
`geom_ridges`, make a plot of traffic over time, within each of the cities. An
example result is shown below.

### (5) Language Learning

This problem will look at a simplified version of the data from the study _A
critical period for second language acquisition: Evidence from 2/3 million
English speakers_, which measured the effect of the the age of initial language
learning on performance in grammar quizzes. We have downloaded the raw data from
the supplementary material and reduced it down to the average and standard
deviations of test scores within (initial learning age) $\times$ (current
age-group) combinations. We have kept a column `n` showing how many participants
were used to compute the associated statistics. The resulting data are available
[here](https://uwmadison.box.com/shared/static/m53dea9w5ipczs3d7nqdnxqxx5ao501b.csv).

a) Using the `.derive()` command in arquero, create two new fields, `low` and
`high`, giving confidence intervals for the means in each row. That is, derive
new variables according to $\hat{x} \pm 2 * \frac{1}{\sqrt{n}}\hat{\sigma}$.
b) Create a `markArea`-based ribbon plot showing confidence intervals for
average test scores as a function of starting age. Include a line for the
average score within that combination. An example result is shown in Figure
\ref{fig:5}. Interpret the results of the study.

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{/Users/kris/Desktop/languages.png}
  \caption{Example result for problem (5).}
  \label{fig:5}
\end{figure}

### (6) Deconstruction

Take a static screenshot from any of the visualizations in this
[article](https://www.theguardian.com/us-news/ng-interactive/2017/dec/20/bussed-out-america-moves-homeless-people-country-study),
and deconstruct its associated visual encodings.

a) What do you think was the underlying data behind the current view? What where
the rows, and what were the columns?
b) What were the data types of each of the columns?
c) What encodings were used? How are properties of marks on the page derived
from the underlying abstract data?
c) Is multi-view composition being used? If so, how?
