---
title: "Principal Components Analysis II"
description: |
    Visualizing and interpreting PCA.
author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: 2021-03-31
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library("knitr")
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_[Reading](),[Recording](),[Rmarkdown]()_

```{r}
library("ggplot2")
library("ggrepel")
library("readr")
library("stringr")
library("tidymodels")
library("tidytext")
theme479 <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    legend.position = "bottom"
  )
theme_set(theme479)
```

1. How should we visualize the results from PCA? There are two artifacts
produced by the procedure — components and scores.

```{r}
# produced by code in previous notes
components <- read_csv("https://uwmadison.box.com/shared/static/dituepd0751qqsims22v2liukuk0v4bf.csv") %>%
  filter(component %in% str_c("PC", 1:5))
scores <- read_csv("https://uwmadison.box.com/shared/static/qisbw1an4lo8naifoxyu4dqv4bfsotcu.csv")
variances <- read_csv("https://uwmadison.box.com/shared/static/ye125xf8800zc5eh3rfeyzszagqkaswf.csv") %>%
  filter(terms == "percent variance")
```

2. First, let's see how much variance is explained by each dimension of the PCA.
Without detouring into the mathematical explanation, know that the more rapid
the dropoff in variance explained, the better the low-dimensional approximation
is. For example, if the data actually lie on a 2D plane in a high-dimensional
space, then the first two bars contain all the variance (the rest would be
zero).

```{r, fig.cap = "Proportion of variance explained by each component in the PCA."}
ggplot(variances) +
  geom_col(aes(component, value))
```


2. Can interpret components by looking at how variables contribute. From the
plot below, we see that the first PC mostly captures variation related to
whether the drink is made with powdered sugar or simple syrup. Drinks with high
values of PC1 are expected to be made from simple syrup, those with low values
of PC1 are likely made from powdered sugar. From the two largest bars in PC2, we
can see that it highlights vermouth vs. nonvermouth.

```{r}
ggplot(components, aes(value, terms, fill = terms)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~component, nrow = 1) +
  labs(y = NULL) +
  theme(axis.text = element_text(size = 7))
```

3. It is often easier read the components if we sort the bars according to their
magnitude. Our usual approach to reordering axes labels, using either `reorder`
or resetting the levels of the defining factor, will reorder all the facts in
the same way. If we want to reorder each facet on its own, we can use the
`reorder_within` function coupled with `scales_*_reorder`, both from the
tidytext package.

```{r, fig.width = 10, fig.height = 6}
components_ <- components %>%
  filter(component %in% str_c("PC", 1:3)) %>%
  mutate(terms = reorder_within(terms, abs(value), component))
ggplot(components_, aes(value, terms, fill = terms)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ component, scales = "free_y") +
  scale_y_reordered() +
  labs(y = NULL) +
  theme(axis.text = element_text(size = 7))
```

4. Next, we can visualize the scores of each sample with respect to these
components. The plot below shows $\left(z_{i1}, z_{i2}\right)$. Suppose that the
columns of $\Phi$ are $\varphi_{1}, \dots, \varphi_{K}$. Then, since $x_{i}
\approx \varphi_{1}z_{i1} + \varphi_{2} z_{i2}$, they have large values for
variables with large component values in the coordinate directions where $z_{i}$
is farther along.

```{r, fig.height = 6}
ggplot(scores, aes(PC1, PC2, label = name)) +
  geom_point(aes(color = category), alpha = 0.7, size = 1.5) +
  geom_text_repel(check_overlap = TRUE, size = 3) +
  coord_fixed(sqrt(variances$value[2] / variances$value[1])) + # rescale axes to reflect variance
  labs(color = NULL)
```

5. For example, El Nino has high value for PC1, which means it has a high value
of variables that are positive for PC1 (like simple syrup) and low value for
those variables that are negative (like powdered sugar).

6. Similarly, since $\varphi_{2}$ puts high positive weight on vermouth-related
variables, H. P. W. Cocktail has many vermouth-related ingredients.

7. In practice, it will often be important to visualize several pairs of PC
dimensions against one another, not just the top 2.

8. Let’s examine the original code in a little more detail. We are using
tidymodels, which is a package for decoupling the definition and the execution
of a data pipeline. This compartmentalization makes it easier to design and
reuse across settings.

```{r}
pca_rec <- recipe(~., data = cocktails_df) %>%
  update_role(name, category, new_role = "id") %>%
  step_normalize(all_predictors()) %>%
  step_pca(all_predictors())

pca_prep <- prep(pca_rec)
```

9. Here is how you would apply PCA without the tidymodels package. You have to
first split the data into the "metadata" that is used to interpret the resulting
scores and the actual numerical variables used by PCA. Then at the end, you have
to join it back in. It's not impossible, but the result is not nearly as
readable.

```{r}
# split name and category out of the data frame
pca_result <- cocktails_df %>%
  select(-name, -category) %>%
  scale() %>%
  princomp()

# join them back into the PCA result
metadata <- cocktails_df %>%
  select(name, category)
scores_direct <- cbind(metadata, pca_result$scores)

ggplot(scores_direct, aes(Comp.1, Comp.2, label = name)) +
  geom_point(aes(color = category), alpha = 0.7, size = 1.5) +
  geom_text_repel(check_overlap = TRUE, size = 3) +
  coord_fixed(sqrt(variances$value[2] / variances$value[1])) + # rescale axes to reflect variance
  labs(color = NULL)
  geom_point()
```

10. The equivalent tidymodels implementation handles the difference between
supplementary and modeling data less bluntly, setting the `name` and `category`
variables to `id` roles, so that `all_predictors()` knows to skip them.

11. We conclude with some characteristics of PCA, which are worth keeping in
mind when choosing between alternative dimensionality reduction methods.

  * *Global structure*: Since PCA is looking for high-variance overall, it tends
  to focus on global structure.
  * *Linear*: PCA can only consider linear functions of the original features.
  If we expect nonlinear features to be more meaningful, then another approach
  should be considered.
  * *Interpretable features*: The PCA components exactly specify how to
  construct each of the derived features.
  * *Fast*: Compared to most dimensionality reduction methods, PCA is quite
  fast. Further, it is easy to implement approximate versions of PCA that scale to
  very large datasets.
  * *Deterministic*: Some embedding algorithms perform an optimization process,
  which means there might be some variation in the results due to randomness in
  the optimization. In contrast, PCA is deterministic, with the components being
  unique up to sign (i.e., you could reflect the components, but the effective
  features will always be the same).
  