---
title: "Introduction to Vega-Lite"
description: |
author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: 01-25-2021
output:
  distill::distill_article:
    self_contained: false
---

You can think of vega-lite as the javascript equivalent of ggplot2. It also
provides a grammar of graphics, where layers can be composed on top of one
another. The main reason for knowing vega-lite (opposed to just using ggplot2
for everything) is that it can support flexible interaction. Also, it’s a good
gateway to understanding richer data visualization packages, like `d3`.

I encourage you to follow along in a new observable
[notebook](https://observablehq.com/@krisrs1128/introduction-to-vega-lite). 

```{r}
library("robservable")
library("htmlwidgets")
```

### Components of a Graph

We’re going to construct this graph, which is from this week's
[reading](https://observablehq.com/@uwdata/data-types-graphical-marks-and-visual-encoding-channels?collection=@uwdata/visualization-curriculum),
step-by-step, just like in the ggplot2 lecture. I realize that this is a bit
repetitive, but I want to make it obvious just how similar ggplot2 and vega-lite
are.

```{r}
robservable("@krisrs1128/introduction-to-vega-lite", include = 4, height = 325)
```

## Data and Geometry

First, let’s load up and preview the
[data](https://uwmadison.box.com/s/dyz0qohqvgake2ghm4ngupbltkzpqb7t). We're
using the arquero library (`aq`) to make it easier to manage data in javascript
-- the filtering and printing steps would be more difficult without it.

````js
data = aq.fromCSV(await FileAttachment("gapminder.csv").text()) // I used the file upload button to add this file
data2000 = data.filter(d => d.year == 2000) // filter to only 2000
data2000.view()
````
```{r}
robservable("@krisrs1128/introduction-to-vega-lite", include = 6, height = 250)
```

and create a basic visual mark, without any encodings. Like in our first
`ggplot2` figure, the mark is collapsed into a single point at the origin.
Notice the analogy betwen vega-lite's `markPoint()` and ggplot2's
`geom_point()`.

````js
vl.markPoint()
  .data(data2000)
  .render()
````

```{r}
robservable("@krisrs1128/introduction-to-vega-lite", include = 7, height = 20)
```

## Encodings

Let’s deconstruct the original figure and see how data fields associated with
each country get mapped to which visual properties of the corresponding point.

* Fertility <--> $x$-axis coordinate
* Life expectancy <--> $y$-axis coordinate
* Country cluster <--> Point color
* Population <--> Point size

We can build this up one step at a time. To encode the fertility and life
expectancy fields, we use `vl.x()` and `vl.y()` in an `encode()` call. One
subtlety is that it's necessary to specify the type of the field through the --
`fieldQ` means that the field is "Quantitative". We'll get into more detail
about specifying data types in the next lecture.

````js
vl.markPoint()
  .data(data2000)
  .encode(
    vl.x().fieldQ("fertility"), // what happens if you change this to fieldN()?
    vl.y().fieldQ("life_expect")
  )
  .render()
````

```{r}
robservable("@krisrs1128/introduction-to-vega-lite", include = 8)
```

The same idea can be used to encode the country cluster and population fields.
Note that I use `fieldN` to specify that cluster is a Nominal variable.

Hint: How do you know which commands encode which fields? Until it becomes
second nature, it's good to keep the
[vega-lite-api](https://vega.github.io/vega-lite-api/api/) and
[vega-lite](https://vega.github.io/vega-lite/) documentation close at hand.

````js
vl.markPoint()
  .data(data2000)
  .encode(
    vl.x().fieldQ("fertility"),
    vl.y().fieldQ("life_expect"),
    vl.color().fieldN("cluster"), // what happens if you change this to fieldO()?
    vl.size().fieldQ("pop")
  )
  .render()
````

```{r}
robservable("@krisrs1128/introduction-to-vega-lite", include = 9)
```

## Finishing touches

We've made headway in our visualization task, but a more critical eye will help
us improve the display. It will also give us an initial look at interactivity.
The main things we ought to improve in this display are,

1. The open rings are a bit distracting. They violate the principle of [Smallest
Effective
Difference](https://sites.google.com/site/tufteondesign/home/practical-design-strategies).
2. Some of the points are impossibly small.
3. The axis labels are still the names from the data.frame.
4. There are distractingly many levels in the legend for population size.
5. We can't actually tell which country each point corresponds to. For this, we
will add a tooltip (a label that appears when you mouseover the point).

For (1), let's replace the open rings with filled circles. We can pass options
to `markPoint` in the same way that we passed options outside of `aes` to ggplot
`geom` layers. For (2), we can modify the default size range by adding on a
`.scale()` to the `vl.size()` encoding -- this is similar to adding a custom
scale in ggplot2.

````js
vl.markPoint({filled: true})
.data(data2000)
.encode(
  vl.x().fieldQ("fertility"),
  vl.y().fieldQ("life_expect"),
  vl.color().fieldN("cluster"),
  vl.size().fieldQ("pop").scale({range: [25, 1000]}) // try changing the ranges
)
.render()
````

```{r}
robservable("@krisrs1128/introduction-to-vega-lite", include = 10)
```

For (3), we can add a `.title()` to fields for which we want to customize
titles, and for (4), we can modify the tick count in the `.legend()` for `size`.

````js
vl.markPoint({filled: true})
.data(data2000)
.encode(
  vl.x().fieldQ("fertility").title("Fertility"),
  vl.y().fieldQ("life_expect").title("Life Expectancy"),
  vl.color().fieldN("cluster"),
  vl.size().fieldQ("pop").scale({range: [25, 1000]}).legend({tickCount: 3}).title("Population"),
)
.render()
````

```{r}
robservable("@krisrs1128/introduction-to-vega-lite", include = 11)
```

Now to (5). The idea is to think of a tooltip as another type of encoding --
just like coordinate positions and colors, a label is a way of transforming a
field in an abstract dataset into a visual property we can observe on a screen.

````js
vl.markPoint({filled: true})
.data(data2000)
.encode(
  vl.x().fieldQ("fertility").title("Fertility"),
  vl.y().fieldQ("life_expect").title("Life Expectancy"),
  vl.color().fieldN("cluster"),
  vl.size().fieldQ("pop").scale({range: [25, 1000]}).legend({tickCount: 3}).title("Population"),
  vl.tooltip().fieldN("country"),
  vl.order().fieldQ('pop').sort('descending') // if you remove this, large points prevent mouseover on the points below them
)
.render()
````

```{r}
robservable("@krisrs1128/introduction-to-vega-lite", include = 12)
```

Finally, for a taste of what's to come, here's a version of the same plot that
let's us compare many years over time.

```{r}
robservable("@krisrs1128/introduction-to-vega-lite", include = 13)
```

Don't worry about the code implementation, but in case you're curious, it's only
a few extra lines over what we've already implemented.

````js
{
  // define a slider selector object
  let selectYear = vl.selectSingle("select").fields("year")
    .init({year: 1955})
    .bind(vl.slider().min(1955).max(2005).step(5).name("Year"))
  
  // update the plot depending on the value of the slider
  return vl.markPoint({filled: true})
    .data(data)
    .select(selectYear) // refers to the slider bar
    .transform(vl.filter(selectYear))
    .encode(
      vl.x().fieldQ("fertility").title("Fertility").scale({domain: [0, 9]}), // hard code the axes ranges
      vl.y().fieldQ("life_expect").title("Life Expectancy").scale({domain: [0, 90]}),
      vl.color().fieldN("cluster"),
      vl.size().fieldQ("pop").scale({domain: [0, 1200000000], range: [25, 1000]}).legend({tickCount: 3}).title("Population"),
      vl.tooltip().fieldN("country"),
      vl.order().fieldQ('pop').sort('descending')
    )
    .render()
}
````
