---
title: ""
description: | 
author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: 2021-04-20
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library("knitr")
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_[Reading](), [Recording](), [Rmarkdown]()_

```{r}
library("dplyr")
library("keras")
library("purrr")
library("RColorBrewer")
```

1. In the last lecture, we discussed the conceptual foundations of feature
learning. In this lecture, we’ll see how to extract and visualize features
learned by a computer vision model.

```{r}
f <- tempfile()
download.file("https://uwmadison.box.com/shared/static/o7t3nt77iv3twizyv7yuwqnca16f9nwi.rda", f)
images <- get(load(f))
length(images) # 20 sample images
```

2. The model at this link was trained to distinguish between photos of cats and
dogs. Let’s take a look at a few examples from the training dataset.

```{r}
images %>%
  array_tree(1) %>%
  map(~ plot(as.raster(., max = 255)))
```

3. The `array_branch` function above splits the 4 dimensional array into a
collection of 3 dimensional slices. Each of these 3D slices corresponds to one
image — the three channels correspond to Red, Green, and Blue colors,
respectively.

4. Next, let’s consider what types of features the model has learned, in order
to distinguish between cats and dogs.

5. We will create an R object to retrieve all the intermediate feature
activations associated with a single image. Every time we call this object on a
new input image, it will return the activations for features at all layers.

```{r}
layer_outputs <- lapply(model$layers, function(layer) layer$output)
activation_model <- keras_model(inputs = model$input, outputs = layer_outputs)
features <- predict(activation_model, images)
```

6. Each element of the list corresponds to a different layer. Within a single
layer, the 3D array of activations provides the activations of each feature
across different spatial windows. For example, for the first layer, there are 32
features, spread over 148 x 148 spatial windows.

```{r}
length(features)
dim(features[[1]])
model
```


7. Let’s visualize a few of these features. Notice that a few of them are edge
detectors — they have high values when there are sharp changes in color in the
underlying images. This is consistent with our earlier claim that the first
layer of a network learns to recognize pixel-level interactions.

```{r}
plot_feature <- function(feature) {
  rotate <- function(x) t(apply(x, 2, rev))
  image(rotate(feature), axes = FALSE, asp = 1, col = brewer.pal(4, "Blues"))
}

ix <- 3
plot(as.raster(images[ix,,, ], max = 255))
plot_feature(features[[1]][ix,,, 1])
```

```{r}
features[[2]][ix,,,1:10] %>%
  array_branch(margin = 3) %>%
  map(~ plot_feature(.))
```

8. Next, we visualize features at a higher level in the network. At this point,
each activation corresponds to a larger spatial context in the original image,
so there are fewer activations per feature (There are more feature maps total,
but each is smaller).

```{r}
plot(as.raster(images[ix,,,], max = 255))
features[[6]][ix,,,1:40] %>%
  array_branch(margin = 3) %>%
  map(~ plot_feature(.))
```

9. It seems that some of the features have learned to recognize higher-level
concepts in the image, like the presence of a cat eye or ear. However, it’s hard
to know definitively, since we are only considering a single image. In the next
set of notes, we will examine the same neuron across many dataset examples, and
this will give us more confidence in our interpretations of individual neurons.

