---
title: "Partial Dependence Profiles I"
description: |
author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: 2021-04-12
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library("knitr")
opts_chunk$set(cache = TRUE, message = FALSE, warning = FALSE, echo = TRUE)
```

_[Reading](), [Recording](), [Rmarkdown]()_

```{r}
library("readr")
library("caret")
library("dplyr")
library("DALEX")
library("ggplot2")
theme479 <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    legend.position = "bottom"
  )
theme_set(theme479)
```

1. As more complex models become more common in practice, visualization has
emerged as a key way for (a) summarizing their essential structure and (b)
motivating further modeling refinements.

2. To ground the discussion, we imagine the standard supervised learning setup.
We approximate the relationship between a $D$-dimensional covariate $x$ and a
univariate response $y$ by a function $f$. We are given a sample of $n$ pairs
$\left(x_{i}, y_{i}\right)$ with which to learn this relationship, and we hope
that the function we learn will generalize to future observations.

3. Some further notation: We will write $x_{j}$ for the $j^{th}$ coordinate of
$x$. We will write $x^{j\vert = z}$ to denote the observation $x$ with the
$j^{th}$ coordinate set to $z$.

4.  Simple linear models are simple enough that they don’t require any follow-up
visual inspection. Since they assume $f\left(x\right) = \hat{\beta}^{T}x$, they
are completely described by the vector of coefficients $\hat{\beta}$. We can
exactly describe what happens to $f$ when we increase $x_{j}$ by one unit: we
just increase the prediction by $\hat{\beta}_{j}$.

5. More complex models — think random forests or neural networks — don’t have
this property. While these models often have superior performance, it’s hard to
say how changes in particular input features will affect the prediction.

6. Partial dependence plots try to address this problem. They provide a
description for how changing the $j^{th}$ input feature affects the predictions
made by complex models.

7. To motivate the definition, consider the toy example below. The surface is
the fitted function $f\left(x\right)$, mapping a two dimensional input $x$ to a
real-valued response. How would you summarize the relationship between $x_{1}$
and $y$? The main problem is that the shape of the relationship depends on which
value of $x_{2}$ we start at.

8. One idea is to consider the values of $x_{2}$ that we actually observed in
our dataset. Then, we can evaluate our model over a range of values $x_{1}$
after fixing those values of $x_{2}$. These curves are called « Ceteris
Paribus » profiles^[Ceteris Paribus means « All Else Held Equal. »].

9. The same principle holds in higher dimensions. We can fix $D - 1$ coordinates
of an observation and then evaluate what happens to a sample’s predictions when
we vary coordinate $j$. Mathematically, this is expressed by $h_{x}^{f,
j}\left(z\right) := f\left(x^{j\vert= z}\right)$.

10. For example, let's consider how CP can be used to understand a model fitted
to a housing prices dataset. Each row of the dataset is a different
neighborhood, and various features about them are collected, like the average
home age and median income of residents. Below, a random forest is used to
predict the median price of homes in these neighborhoods.

```{r}
data(titanic)
titanic <- titanic %>%
  select(-country) %>%
  na.omit()

x <- select(titanic, -survived)
hyper <- data.frame(n.trees = 100, interaction.depth = 8, shrinkage = 0.1, n.minobsinnode = 10)
fit <- train(x = x, y = titanic$survived, method = "gbm", tuneGrid = hyper, verbose = F)
```

11. Next, we can compute the CP profiles. We are showing median income, though
any subset of variables could be requested. The bold curve is a Partial
Dependence (PD) profile, which we will discuss below. Each of the other curves
corresponds to a neighborhood, though only a subsample of neighborhoods are
shown. The curves are obtained by fixing all the features of a neighborhood
except for the median income, and then seeing what happens to the prediction
when the median income feature is increased or decreased.

```{r, fig.cap = "CP and PDP profiles for median income, for a random forest fitted to the California housing prices dataset."}
explanation <- explain(model = fit, data = x, y = titanic$survived)
profile <- model_profile(explainer = explanation)
plot(profile, geom = "profiles", variables = "age") +
  theme479
```

12. It seems that, up to a certain income, the median income is linearly related
to predicted median home price. Beyond this point, the predictions flatten out.
Beyond that point, there are likely too few observations to make a good
prediction, so the model fixes its value for all neighborhoods with larger
incomes. Notice that the profiles seem to vertically offset from one
neighborhood to another, this is because, aside from median income, each
neighborhood has features that make it more or less likely to have high median
home prices.

13. We used the DALEX package to produce these curves. The `explain` function
takes the fitted model and original dataset as input. It returns an object with
many kinds of model summaries. To extract the CP profiles from these summaries,
we use `model_profile`. The output of this function has been designed so that
calling `plot` with `geom = "profiles"` will show the CP profiles.

14. The PD profile is computed by averaging across all the CP profiles. It is a
more concise alternative to CP profiles, showing one curve per features, rather
than one curve per sample.

```{r, fig.width = 10, fig.height = 5}
plot(profile, geom = "aggregates") +
  theme479
```

14. Not only are the PD plots simpler to read than the full collection of CP
profiles — by performing this aggregation step, subtle patterns may become more
salient, for the same reason that an average carries more information than any
subset of observations.

```{r}
save(fit, file = "~/Desktop/fit.rda")
```

