---
title: ""
description: |

author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: 2021-04-04
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library("knitr")
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_Reading [1](), [Recording](), [Rmarkdown]()_

```{r}
library("ggplot2")
library("dplyr")
library("forcats")
library("tidyr")
library("topicmodels")
library("superheat")
library("readr")
library("tidytext")
theme479 <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    legend.position = "bottom"
  )
theme_set(theme479)
```

1. We have used text data analysis to motivate and illustrate the use of topic
models. However, these models can be used whenever we have high-dimensional
count data. To illustrate this broad applicability, this lecture will consider
an example from gene expression analysis.

2. The dataset we consider comes from the GTEX consortium. A variety of tissue
samples have been subject to RNA-seq analysis, which measures how much of each
type of gene is expressed within each sample. Intuitively, we relate,

	* Documents $\arrow$ Tissue samples
	* Words  $\arrow$ Genes
	* Word Counts $\arrow$ Gene expression levels
	
The code below reads in the expression data.

```{r}
x <- read_csv("~/Desktop/gtex.csv")
x_dtm <- cast_dtm(x, sample, gene, value)
#fit <- LDA(x_dtm, k = 10, control = list(seed = 479))
#save(fit, file = "~/Desktop/lda_gtex.rda")
fit <- get(load("~/Desktop/lda_gtex.rda"))
```
	
3. The goal here is like in the cluster stability lecture earlier: we want to
find sets of genes that tend to be expressed together, because they be
indications of some underlying biological processes. Unlike clustering, which
assumes each sample is described by one gene expression profile, a topic model
will be able to model each tissue sample as a mixture of profiles (i.e., a
mixture of underlying biological processes).

4. As a first step in our analysis, we need to prepare a `DocumentTermMatrix`
for use by the topicmodels package. This is in fact more straightforward for
this application than for the text example, since we don’t need to do any text
parsing or pivoting. The data come to us as a wide count matrix.

5. Next, we visualize the topics. Let’s consider the genes with the highest
discrimination between topics. These are genes that are expressed highly in some
topics, but not others. (interpret)

```{r}
tissue_info <- x %>%
  select(sample, starts_with("tissue"))

topics <- tidy(fit, matrix = "beta") %>%
  mutate(topic = factor(topic))
memberships <- tidy(fit, matrix = "gamma") %>%
  mutate(topic = factor(topic)) %>%
  left_join(tissue_info, by = c("document" = "sample"))
```

```{r, echo = FALSE}
kl_div <- function(p1, p2) {
  p1 * log(p1 / p2) + (p2 - p2)
}

kl_mat <- function(p) {
  K <- matrix(0, nrow = length(p), ncol = length(p))
  
  for (i in seq_along(p)) {
    for (j in seq_len(i - 1)) {
      K[i, j] <- kl_div(p[i], p[j])
    }
  }
  K
}

discrepancy <- function(p, lambda = 1e-7) {
  p <- (p + lambda) / sum(p + lambda) # Laplace smoothing
  K <- kl_mat(p)
  max(K)
}
```

```{r, fig.height = 8, fig.width = 4.5}
discriminative_genes <- topics %>%
  group_by(term) %>%
  mutate(D = discrepancy(beta)) %>%
  ungroup() %>%
  slice_max(D, n = 400) %>%
  mutate(term = fct_reorder(term, -D))

discriminative_genes %>%
  pivot_wider(term, topic, values_from = "beta") %>%
  column_to_rownames("term") %>%
  superheat(
    pretty.order.rows = TRUE,
    left.label.size = 1.5,
    left.label.text.size = 3,
    bottom.label.size = 0.05,
    legend = FALSE
  )
```

6. Now, let’s see what tissues express which topics. We can use a structure
plot.

```{r}
```

```{r}
hclust_result <- hclust(dist(fit@gamma))
document_order <- fit@documents[hclust_result$order]
memberships <- memberships %>%
  mutate(document = factor(document, levels = document_order))
```

```{r, fig.height = 10, fig.width = 5}
keep_tissues <- memberships %>%
  count(tissue) %>%
  filter(n > 80) %>%
  pull(tissue)

memberships %>%
  filter(tissue %in% keep_tissues) %>%
  ggplot(aes(gamma, document, fill = topic, col = topic)) +
  geom_col(position = position_stack()) +
  facet_grid(tissue ~ ., scales = "free", space = "free") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_color_brewer(palette = "Set3") +
  scale_fill_brewer(palette = "Set3") +
  theme(
    panel.spacing = unit(0.5, "lines"),
    strip.switch.pad.grid = unit(0, "cm"),
    strip.text.y = element_text(size = 8, angle = 0),
    axis.text.y = element_blank(),
  )
```

7. (interpret)