---
title: "Spatial Data Interaction"
description: |
  Some strategies for interactively visualizing spatial data.
author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: 03-12-2021
output:
  distill::distill_article:
    self_contained: false
---

_[Reading](https://geocompr.robinlovelace.net/spatial-class.html), [Recording](https://mediaspace.wisc.edu/media/Week%207%20%5B5%5D%20Spatial%20Data%20Interaction/1_9ccr93kw), [Rmarkdown](https://github.com/krisrs1128/stat479/blob/master/_posts/2021-03-02-week7-5/week7-5.Rmd)_

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

```{r}
library("robservable")
library("leaflet")
library("readr")
```

1. These are optional notes. The reading does not cover interactivity in spatial
data contexts, and there will be no problems in this class related to
interactive spatial data visualization. However, these notes may be useful if
your project involves interaction with spatial data.

2. In `vega-lite`, vector data can be visualized using `vl.markGeoshape`. Any
fields contained within the vector dataset can be visually encoded in the marks.
This is analogous to `geom_sf` in ggplot2.

````js
vl.markGeoshape({fill: "#e0e0e0", stroke: "#f7f7f7", strokeWidth: 1})
    .data(vector_data) // contains the vector data boundaries
    .project(vl.projection('albersUsa'))
    .render()
````

```{r}
robservable("@krisrs1128/minimal-geoshape", 3, height = 300)
```

3. Spatial data are often amenable to dynamic linking. For example, selections
in a complementary figure can be used to query points on a map. Here are two
examples from public observable notebooks. The first, by Chanwut Kittivorawong,
updates the cities displayed on a map using a brush applied to histogram data.
Conceptually, it is like the NYC house prices homework problem, except it
includes several histogram selectors and draws a background map for context.

```{r}
robservable("@chanwutk/happiness-of-each-country", 6, height = 750)
```
The second, by Alissa Chen, shows how the amount of daylight varies across
cities, depending on time of the year. Brushing over the time series filters the
data that are used to determine the circle widths (the little suns) on top of
the map.

```{r}
robservable("@alchan/whats-the-average-amount-of-sunlight-in-a-city-over-a-short-te", 3, height = 730)
```


4. Alternatively, selections on the map can highlight elements in a
complementary figure. For example, hovering over geographic elements might
highlight series in a time series plot or update a barchart placed on the side.
Vega-lite doesn't support selections based on vector data elements, only points
that are overlaid on the map. However, these sorts of selections can be made
with d3.

```{r}
robservable("@krisrs1128/geojson-mouseover", 7)
```

5. Sometimes it’s useful to add an interactive map background to a vector
dataset. In this case, the `leaflet` package in R can be useful. In the code
block below, `addTiles` fetches the background map. `addCircles` overlays the
new vector features on top of the map. It's worth noting that the vector
features were created automatically -- there was no need to create or read in
any type of `sf` object.

```{r}
cities <- read_csv("https://uwmadison.box.com/shared/static/j98anvdoasfb1h651qxzrow2ua45oap1.csv")
leaflet(cities) %>% 
  addTiles() %>% 
  addCircles(
    lng = ~Long, 
    lat = ~Lat, 
    radius = ~sqrt(Pop) * 30
  )
```

