---
title: ""
description: |

author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: 2021-04-04
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library("knitr")
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_Reading [1](), [Recording](), [Rmarkdown]()_

```{r}
```

1. Topic modeling is a type of dimensionality reduction method that is
especially useful for high-dimensional count matrices.

2. In text data analysis, each row is a document and each column is a word. The
$ij^{th}$ entry contains the count of the $j^{th}$ word in the $i^{th}$ document.

3. In gene expression analysis, each row is a biological sample, and each column
is a gene. The $ij^{th}$ entry contains the count of the number of sequencing
reads in the $i^{th}$ sample that align with the the $j^{th}$ gene.

4. For clarity, we will refer to samples as documents and features as words.
However, keep in mind that these methods are in fact more general.

5. For these type of data, topic models provide a compromise between clustering
and PCA.

6. In clustering, each document would have to be assigned to a single topic. In
contrast, topic models allow each document to partially belong to several topics
simultaneously. In this sense, they are more suitable to data that varies
continuously along a gradient.

7. In PCA, there is no notion of clusters. In contrast, topic models do estimate
$K$ topics, which are analogous to a cluster centroids (though documents are
usually a mix of several centroids).

8. Without delving into the mathematical detail, topic models perform
dimensionality reduction by supposing,

* Each document is a mixture of topics.
* Each topic is a mixture of words.

5. To understand the first point, consider modeling a collection of newspaper
articles. A set of articles might belong primarily to the « politics » topic,
and others to the « business » topic. Articles that describe a monetary policy
in the federal reserve might belong partially to both the « politics » and the
« business » topic.

6. To understand the second point, consider the difference in words that would
appear in politics and business articles. Articles about politics might
frequently include words like « congress » and « law », but only rarely words
like « stock » and « NYSE. »

7. Geometrically, LDA can be represented by the following picture. The corners
of this simplex represent different words (in reality, there would be $V$
different corners to this simplex). A topic is a point on this simplex. The
closer the topic is to one of the corners, the more frequently that word appears
in the topic.

8. A document is a point intermediate between a set of topics. A document that’s
very close to a particular topic has a word distribution just like that topic. A
document that’s intermediate between two topics has a word distribution that
mixes between both topics