---
title: ""
description: |

author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: 2021-04-04
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library("knitr")
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_Reading [1](), [Recording](), [Rmarkdown]()_

```{r}
library("tidytext")
library("topicmodels")
library("stringr")
library("gutenbergr")
library("dplyr")
library("ggplot2")
library("tidyr")
theme479 <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    legend.position = "bottom"
  )
theme_set(theme479)
```

1. There are several packages in R that can be used to fit topic models. We will
use LDA as implemented in the `topicmodels` package,  which expects input to be
structured as a `DocumentTermMatrix`, a special type of matrix that stores the
counts of words (columns) across documents (rows). In practice, most of the
effort required to fit a topic model goes into transforming the raw data into a
suitable `DocumentTermMatrix`.

2. To illustrate this process, let’s consider the « Great Library Heist »
example from the reading. We imagine that a thief has taken four books — Great
Expectations, Twenty Thousand Leagues Under The Sea, War of the Worlds, and
Pride & Prejudice — and torn all the chapters out. We are left with pieces of
isolated pieces of text and have to determine from which book they are from.

```{r}
titles <- c("Twenty Thousand Leagues under the Sea", 
            "The War of the Worlds",
            "Pride and Prejudice", 
            "Great Expectations")
books <- gutenberg_works(title %in% titles) %>%
  gutenberg_download(meta_fields = "title")
books
```
3. Since we imagine that the word distributions are not equal across the books,
topic modeling is a reasonable candidate for recovering this information. Note
that, in principle, other clustering and dimensionality reduction procedures
could also work.

4. First, let’s simulate the process of tearing the chapters out. The objects
below contain the raw text of the four books. We split the raw texts anytime the
word « Chapter » appears. We will keep track of the book names for each chapter,
but note that it is not passed into the topic modeling routine.
```{r}
books %>%
  filter(title == "Pride and Prejudice") %>%
  mutate(
    chapter = cumsum(str_detect(text, regex("chapter [0-9]+", ignore_case = TRUE)))
  ) %>%
  group_by(chapter) %>%
  filter(n() > 10)
```

```{r}
# divide into documents, each representing one chapter
by_chapter <- books %>%
  group_by(title) %>%
  mutate(
    chapter = cumsum(str_detect(text, regex("chapter [0-9]+", ignore_case = TRUE)))
  ) %>%
  group_by(title, chapter) %>%
  mutate(n = n()) %>%
  filter(n > 5) %>%
  ungroup() %>%
  unite(document, title, chapter)

# find document-word counts
word_counts <- by_chapter %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words) %>%
  group_by(document, word) %>%
  summarise(n = n())
```

5. As it is, the text data are long character strings, giving actual text from
the novels. LDA throws away information related to word order — it just wants
counts of each word within each each chapter. To derive word counts, we first
split the raw text into separate words using the tidytext package. Then, we can
count the number of times each word appeared in each document using `count`, a
shortcut for the usual `group_by` and `summarize(n = n())` pattern.

```{r}
word_counts <- by_chapter %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words) %>%
  count(document, word)

word_counts
```

6. These wordcounts are still not in a format compatible with conversion to a
`DocumentTermMatrix`. The issue is that the `DocumentTermMatrix` expects words
to be arranged along columns, but currently they are stored across rows. The
line below converts the original « long » word counts dataset into a
`DocumentTermMatrix` in one step.

```{r}
chapters_dtm <- word_counts %>%
  cast_dtm(document, word, n)
chapters_dtm
```

8. Once the data are in this format, we can use the `LDA` function to fit a
topic model. Various hyperparameters can be set using the `control` argument.

```{r}
chapters_lda <- LDA(chapters_dtm, k = 4, control = list(seed = 1234))
chapters_lda
```

9. There are two types of outputs produced by the LDA model: the topic word
distributions (for each topic, which words are common?) and the document-topic
memberships (for which topics is does a document come from?). For visualization,
it will be easiest to extract these parameters using the `tidy` function,
specifying whether we want the topics (beta) or memberships (gamma).

```{r}
topics <- tidy(chapters_lda, matrix = "beta")
memberships <- tidy(chapters_lda, matrix = "gamma")
```

10. This tidy approach is preferable to extracting the parameters directly
because it ensures the output is a tidy data.frame, rather than a matrix, and
tidy data.frames are easier to visualize in ggplot2.

```{r}
# highest weight words per topic
topics %>%
  arrange(topic, -beta)

# topic memberships per document
memberships %>%
  arrange(document, topic)
```

