---
title: ""
description: |

author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: 2021-04-04
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library("knitr")
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

```{r}
library("tidytext")
library("topicmodels")
library("stringr")
library("tibble")
library("forcats")
library("readr")
library("ggridges")
library("dplyr")
library("ggplot2")
library("superheat")
library("tidyr")
theme479 <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    legend.position = "bottom"
  )
theme_set(theme479)
```

1. In the last set of notes, we fit and extracted parameters from a topic model
applied to the « Great Library Heist » dataset, but we did not visualize or
interpret the results. We’ll continue that here, with eye towards the types of
visualizations that might be generally useful whenever a topic model is fit.

```{r}
memberships <- read_csv("~/Desktop/memberships.csv")
topics <- read_csv("~/Desktop/topics.csv")
```

2. A topic is a probability distribution across a collection of words. If the
vocabulary isn’t too large, two appropriate visualization strategieis are,

	* Faceted barplot: Each facet corresponds to a topic. The height of each bar
	corresponds to a given word’s probability within the topic. The sum of heights
	across all bars is 1. 
	* Heatmap: Each row is a topic and each color is a word.
	The color of the heatmap cells gives the probability of that word within the
	given topic.
	
3. We can construct a faceted barplot using the tidied beta matrix.

```{r}
ggplot(topics %>% filter(beta > 3e-4), aes(term, beta)) +
  geom_col() +
  facet_grid(topic ~ .) +
  theme(axis.text.x = element_blank())
```

4. For the heatmap, we need to pivot the topics, so that words appear along
columns. From there, we can use superheatmap.

```{r}
topics %>%
  filter(beta > 3e-4) %>%
  pivot_wider(topic, term, values_from = "beta", values_fill = 0, names_repair = "unique") %>%
  select(-1) %>%
  superheat(pretty.order.cols = TRUE)
```

5. Both approaches breakdown when the vocabulary is very large — there are
simply too many words to include a label for each. A workaround is to restrict
attention to a subset of « interesting » words. For example, we could filter to,

  * Top words overall: We can filter to words that appear above some threshold
  across any topic. This is the approach used in the visualizations above,
  though the threshold is very low (there are still too many words to add
  labels).
	* Top words per topic: We can sort the words within each topic in order from
	highest to lowest probability, and then keep only the $S$ largest.
	* Most discriminative words: Some words have high probability just because they
	are common. They have high probability within each topic, and so are not truly
	that interesting. Instead, we can focus on words that have very different
	probabilities across the topics. For example, we can use the KL divergence,
	
6. To visualize the top words, we first filter the topic matrix. Then, we use
the same `reorder_within` function from the PCA lectures to reorder words within
each topic.

```{r}
top_terms <- topics %>%
  group_by(topic) %>%
  slice_max(beta, n = 10) %>%
  mutate(term = reorder_within(term, beta, topic))
ggplot(top_terms, aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free") +
  scale_y_reordered()
```

7. To visualize discriminative words, we first compute the discrimination
measure for each word and filter to those with the top score. The filtered
results can be used in either the faceted barplot or heatmaps above.

```{r, echo = FALSE}
kl_div <- function(p1, p2) {
  p1 * log(p1 / p2) + (p2 - p2)
}

kl_mat <- function(p) {
  K <- matrix(0, nrow = length(p), ncol = length(p))
  
  for (i in seq_along(p)) {
    for (j in seq_len(i - 1)) {
      K[i, j] <- kl_div(p[i], p[j])
    }
  }
  K
}

discrepancy <- function(p, lambda = 1e-7) {
  p <- (p + lambda) / sum(p + lambda) # Laplace smoothing
  K <- kl_mat(p)
  max(K)
}
```


```{r}
discriminative_terms <- topics %>%
  group_by(term) %>%
  mutate(D = discrepancy(beta)) %>%
  ungroup() %>%
  slice_max(D, n = 200) %>%
  mutate(term = fct_reorder(term, -D))

ggplot(discriminative_terms, aes(term, beta)) +
  geom_col() +
  facet_grid(topic ~ ., scales = "free") +
  theme(axis.text.x = element_text(angle = 90))
```
```{r, fig.height = 7, fig.width = 3}
discriminative_terms %>%
  pivot_wider(term, topic, values_from = "beta") %>%
  column_to_rownames("term") %>%
  superheat(
    pretty.order.rows = TRUE,
    left.label.size = 1.5,
    left.label.text.size = 3,
    bottom.label.size = 0.05,
    legend = FALSE
  )
```

8. Both visualization approaches make it clear that the topics are recovering
the book identities.

9. Besides the topics, we are interested in knowing the topic proportions for
each chapter. One compact approach is to visualize the proportions as a boxplot.

```{r}
memberships <- memberships %>%
  mutate(
    book = str_extract(document, "[^_]+"),
    topic = factor(topic)
  )

ggplot(memberships, aes(topic, gamma)) +
  geom_boxplot()

ggplot(memberships, aes(topic, gamma)) +
  geom_boxplot() +
  facet_wrap(~book)
```

10. An equivalent visualization is to use ridgeplots.

```{r}
ggplot(memberships, aes(gamma, topic)) +
  geom_density_ridges(bandwidth = 0.01)

ggplot(memberships, aes(gamma, topic)) +
  geom_density_ridges(aes(fill = book), bandwidth = 0.01, alpha = 0.8)
```

11. Both approaches above consider the collection of documents in aggregate. If
we want to avoid aggregation and visualize individual documents, we can use a
heatmap or jittered scatterplot. These approaches are useful here because
heatmap cells and individual points can be drawn very small — anything requiring
more space would become unwieldy as the number of documents grows.

```{r, fig.height = 10, fig.width = 3}
memberships_wider <- memberships %>%
  pivot_wider(document:book, topic, values_from = "gamma")

superheat(
  memberships_wider %>% column_to_rownames("document") %>% select(-book),
  pretty.order.rows = TRUE,
  bottom.label.size = 0.05,
  legend = FALSE
)
```


```{r, fig.height = 4, fig.width = 6}
ggplot(memberships, aes(topic, gamma, col = book)) +
  geom_point(position = position_jitter(h = 0.05, w = 0.3))
```

12. Alternatively, we can use a « structure » plot. This is a type of stacked
barplot where the colors of each bar corresponds to a topic. The documents can
be sorted according to either the proportions of each topic or using an a
hierarchical clustering, like in superheatmap’s `pretty.order.*` functions.

```{r, fig.height = 10, fig.width = 4}
document_order <- memberships_wider %>%
  arrange(book, `1`, `2`, `3`, `4`) %>%
  pull(document)
memberships <- memberships %>%
  mutate(document = factor(document, levels = document_order))

ggplot(memberships, aes(gamma, document, fill = topic, col = topic)) +
  geom_col(position = position_stack()) +
  facet_grid(book ~ ., scales = "free", space = "free") +
  scale_x_continuous(expand = c(0, 0)) +
  theme(axis.text.y = element_blank())
```