---
title: ""
description: |

author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: 
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
library("robservable")
```

_[Reading](), [Recording](), [Rmarkdown]()_

1. In $K$-means, $K$ controls the « magnification » of a clustering. What if we
wanted to do the clustering at many different scales?

2. A tree can be used to implicitly store many clusterings at once. You can get
a standard clustering by « cutting » the tree at some level.

3. These hierarchical clustering trees can be thought of abstract versions of
the taxonomic trees. Instead of relating species, they relate observations in a
dataset.

```{r}
robservable("@mbostock/tree-of-life")
```

4. Elaborating on this analogy, the leaves of a hierarchical clustering tree are
the original observations. The more recently two nodes share a common
« ancestor, » the more similar those observations are.

5. The specific algorithm proceeds as follows,

	* Initialize: Associate each point with a cluster \def\reals{\mathbb{R}}
	\def\E{\mathbb{E}} % Expectation symbol \def\P{\mathbb{P}} % Expectation symbol
	\def\Earg#1{\E\left[{#1}\right]} \def\Parg#1{\P\left[{#1}\right]}
	\def\Esubarg#1#2{\E\_{#1}\left[{#2}\right]} \def\eps{\epsilon}
	\def\F{\mathcal{F}} C\_i := \{x\_i\}Ci :={x i }
	* Iterate until only one cluster: Look at all pairs of clusters. Merge the pair
	\def\reals{\mathbb{R}} \def\E{\mathbb{E}} % Expectation symbol
	\def\P{\mathbb{P}} % Expectation symbol \def\Earg#1{\E\left[{#1}\right]}
	\def\Parg#1{\P\left[{#1}\right]} \def\Esubarg#1#2{\E_{#1}\left[{#2}\right]}
	\def\eps{\epsilon} \def\F{\mathcal{F}} C_k, C_{k^{\prime}}C k​	 ,C k ′  which
	are the most similar.\_
	
6. In R, this can be accomplished by using the `hclust` function. First, we need
to compute the distances between all pairs of observations (this provides the
similarities used in the algorithm). Then, we apply `hclust` to the matrix of
pairwise distances.

7. We apply this to a movie ratings dataset. Movies are considered similar if
they tend to receive similar ratings across all audience members. The result is
visualized below.

8. We can cut the tree to recover a standard clustering.

