---
title: "A Vocabulary of Marks"
description: | 
  Examples of marks and their encodings in both ggplot2 and vega-lite.
author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: 01-29-2021
output:
  distill::distill_article:
    self_contained: false
---

Reading: [Data Types, Graphical Marks, and Visual Encoding Channels](https://observablehq.com/@uwdata/data-types-graphical-marks-and-visual-encoding-channels?collection=@uwdata/visualization-curriculum)

* [Recording]
* [Rmarkdown](https://raw.githubusercontent.com/krisrs1128/stat479/master/_posts/2021-01-18-week1-5/week1-5.Rmd)
* [Observable](https://observablehq.com/@krisrs1128/vocabulary-of-marks)

## Vocabulary of encodings

With this in mind, it can only ever help you to have a rich vocabulary of
possible visual encodings. The more kinds of marks and encodings that are at
your fingertips, the better your chances are that you’ll arrive at a
configuration that helps you achieve your purpose.

So, let’s look at a few different types of marks and encodings in R and
vega-lite.

### Point Marks

Point marks can encode data fields using their $x$ and $y$ positions, color,
size, and shape.

```{r}
library("readr")
library("ggplot2")
library("dplyr")
library("robservable")
theme_set(theme_bw())
gapminder <- read_csv("https://uwmadison.box.com/shared/static/dyz0qohqvgake2ghm4ngupbltkzpqb7t.csv", col_types = cols()) %>%
  mutate(cluster = as.factor(cluster)) # specify that cluster is nominal
```

```{r}
gap2000 <- gapminder %>%
  filter(year == 2000) # keep only year 2000

ggplot(gap2000) +
  geom_point(aes(x = fertility, y = cluster, shape = cluster))
```

The equivalent in vega-lite is
````js
vl.markPoint()
  .data(data2000)
  .encode(
    vl.x().fieldQ('fertility'),
    vl.y().fieldN('cluster'),
    vl.shape().fieldN('cluster')
  )
  .render()
````

We can specify different types of shapes using the "shape" parameter outside of
the `aes` encoding.

```{r}
ggplot(gap2000) +
  geom_point(aes(x = fertility, y = cluster), shape = 15)
```

````js
vl.markSquare({size: 100})
  .data(data2000)
  .encode(
    vl.x().fieldQ('fertility'),
    vl.y().fieldN('cluster')
  )
  .render()
````

### Bar Marks

Bar marks let us associate a continuous field with a nominal one.

```{r}
ggplot(gap2000) +
  geom_bar(aes(x = country, y = pop), stat = "identity") +
  theme(
    axis.text.x = element_text(angle = 90),
    panel.grid.major.x = element_blank()
  )
```

````js
vl.markBar()
  .data(data2000)
  .encode(
    vl.x().fieldN('country'),
    vl.y().fieldQ('pop')
  )
  .render()
````

To make comparisons between countries with similar populations easier, we can
order them by population. To compare clusters, we can shade the bars.

```{r}
ggplot(gap2000) +
  geom_bar(
    aes(x = reorder(country, -pop, mean), y = pop, fill = cluster), 
    stat = "identity"
  ) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(expand = c(0, 0, .1, .1)) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.2, size = 8), # more readable axes
    panel.grid.major.x = element_blank()
  )
```

````js
vl.markBar()
  .data(data2000)
  .encode(
    vl.x().fieldN('country').sort("-y"),
    vl.y().fieldQ('pop'),
    vl.color().field("cluster").scale({"scheme": "set2"})
  )
  .width(700)
  .render()
````

In the plot above, each bar is anchored at 0. Instead, we could have each
rectangle encode *two* continuous values, a start and an end. To illustrate,
let's compare the minimum and maximimum life expectancies within each country
cluster.

`vega-lite` let's you compute the summary statistics within the encoding step.
For ggplot2, we'll need to put the summary into its own data frame. For this, we
`group_by` each cluster, so that a `summarise` call finds the minimum and
maximum life expectancies restricted to each cluster.

```{r}
# find summary statistics
life_ranges <- gap2000 %>%
  group_by(cluster) %>%
  summarise(
    min_life = min(life_expect),
    max_life = max(life_expect)
  )

# look at a few rows
head(life_ranges)

# plot them
ggplot(life_ranges) +
  geom_segment(
    aes(x = min_life, xend = max_life, y = cluster, yend = cluster),
    size = 4
  ) +
  xlim(0, 85)
```


````js
vl.markBar()
  .data(data2000)
  .encode(
    vl.x().min('life_expect'),
    vl.x2().max('life_expect'),
    vl.y().fieldN('cluster')
  )
  .render()
````

### Line Marks

Line marks are useful for comparing changes. Our eyes naturally focus on rates
of change when we see lines. Below, we'll plot the fertility over time, colored
in by country cluster.

```{r}
ggplot(gapminder) +
  geom_line(
    aes(x = year, y = fertility, col = cluster, group = country),
      alpha = 0.7, size = 0.9
  ) +
  scale_color_brewer(palette = "Set2")
```

````js
vl.markLine({opacity: 0.7, size: 3})
  .data(data)
  .encode(
    vl.x().fieldO('year'),
    vl.y().fieldQ('fertility'),
    vl.color().fieldN('cluster').scale({"scheme": "set2"}),
    vl.detail().fieldN("country"),
    vl.tooltip().fieldN('country')
  )
  .width(500)
  .render()
````


### Area Marks

Area marks have a flavor of both bar and line marks. The filled area supports
absolute comparisons, while the changes in shape suggest derivatives.

```{r}
population_sums <- gapminder %>%
  group_by(year, cluster) %>%
  summarise(total_pop = sum(pop))

head(population_sums)

ggplot(population_sums) +
  geom_area(
    aes(x = year, y = total_pop, fill = cluster)
  ) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(expand = c(0, 0, .1, .1)) +
  scale_x_continuous(expand = c(0, 0))
```

````js
vl.markArea()
  .data(data)
  .encode(
    vl.x().fieldO('year'),
    vl.y().sum('pop'),
    vl.color().fieldN('cluster').scale({"scheme": "set2"}),
    vl.detail().fieldN("country"),
    vl.tooltip().fieldN('country')
  )
  .width(600)
  .render()
````

Just like in bar marks, we don't necessarily need to anchor the $y$-axis at 0.
For example, here the bottom and top of each area mark is given by the 30 and
70% quartiles of population within each country cluster.

```{r}
population_ranges <- gapminder %>%
  group_by(year, cluster) %>%
  summarise(min_pop = quantile(pop, 0.3), max_pop = quantile(pop, 0.7))

head(population_ranges)

ggplot(population_ranges) +
  geom_ribbon(
    aes(x = year, ymin = min_pop, ymax = max_pop, fill = cluster),
    alpha = 0.8
  ) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(expand = c(0, 0, .1, .1)) +
  scale_x_continuous(expand = c(0, 0))
```

````js
population_ranges = data.groupby("year", "cluster")
  .rollup({ // equivalent of dplyr summarise
    pop_low: op.quantile("pop", 0.4),
    pop_high: op.quantile("pop", 0.7)
  })
````

````js
vl.markArea({"opacity": 0.7})
  .data(population_ranges)
  .encode(
    vl.x().fieldO('year'),
    vl.y().q1('pop_low').title("Population"),
    vl.y2().q3('pop_high'),
    vl.color().fieldN('cluster').scale({"scheme": "set2"})
  )
  .width(600)
  .render()
````