---
title: "Visualizing Convolutional Filters"
description: | 
author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: 2021-04-22
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library("knitr")
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_[Reading](), [Recording](), [Rmarkdown]()_

```{r}
library("dplyr")
library("purrr")
library("keras")
library("tensorflow")
```

1. So far, we’ve visualized neural networks by analyzing the activations of
learned features across observed samples. A complementary approach is to ask
instead — is there a hypothetical image that would maximize the activation of a
particular neuron? If we can construct such an image, then we might have a
better sense of the types of image concepts that a neuron is highly sensitive
to.

2. We will illustrate these ideas on a network that has been trained on
Imagenet. This is a large image dataset with many (thousands) of class labels,
and it is often used to evaluate image classification algorithms. The network is
loaded below and a summary of its layers is printed.

```{r}
model <- application_vgg16(
  weights = "imagenet", 
  include_top = FALSE
)
activation_model <- keras_model(inputs = model$input, outputs = model$layers[[8]]$output)
```


```{r}
mean_activation <- function(image, layer, filter_ix=1) {
  h <- layer(image)
  k_mean(h[,,, filter_ix])
}

squash <- function(im) {
  (im - min(im)) / (max(im) - min(im))
}

gradient_step <- function(image, layer, filter_ix=1, lr=1e-3) {
  with(tf$GradientTape() %as% tape, {
    tape$watch(image)
    objective <- mean_activation(image, layer, filter_ix)
  })
  grad <- tape$gradient(objective, image)
  image <- image + lr * grad
}
  
random_image <- function() {
  tf$random$uniform(map(c(1, 150, 150, 3), as.integer))
}

visualize_filter <- function(layer, filter_ix = 1, n_iter = 100, lr = 10) {
  im_seq <- array(0, dim = c(n_iter, 150, 150, 3))
  image <- random_image()
  for (i in seq_len(n_iter)) {
    image <- gradient_step(image, layer, filter_ix, lr)
    im_seq[i,,,] <- as.array(image[1,,,])
  }
  
  im_seq
}
```


```{r}
image <- random_image()
updated_image <- gradient_step(image, activation_model)
```

```{r, fig.show = "animate"}
activation_model <- keras_model(inputs = model$input, outputs = model$layers[[3]]$output)
im_seq <- visualize_filter(activation_model)
for (i in seq_len(100)) {
  plot(as.raster(squash(im_seq[i,,,])))
}
```
3. The main idea is to setup an optimization problem that searches through image
space for an image that maximies the activation for a particular neuron.

4. To implement this, we can compute the gradient of a neuron’s activation with
respect to image pixel values. This is a measure of how much the activation
would change when individual pixel values are perturbed.

	1. (draw a picture)
	2. (give code example)
	
5. Once these gradients can be computed, it’s possible to perform gradient
ascent to solve the activation maximization problem.

	1. (draw a picture)
	2. (give code example)
	
6. We can now visualize the image that would have optimized the activation for
neuron 1 in layer 1. It seems to be most responsive to. A look at a few other
neurons at this layer suggest that this layers has learned to recognize …
patterns.

7. We can think of these features as analogous to a collection of basis
functions. At the first layer, the network is representing each image as a
combination of « basis » images, related to particular edge or color patterns.

8. We can compare these activation maximizing inputs with those associated with
later layers. It seems that the basis images at this level are more intricate,
reflecting textures and common objects across this dataset.
