---
title: "SHAP Computation"
author: Kris Sankaran
date: 2026-03-02
bibliography: references.bib
format:
  html:
    embed-resources: true
    html-math-method: mathjax
execute:
    echo: true
    message: false
    warning: false
    cache: false
    eval: false
---

::: {style="display: none;"}
$$
\newcommand{\bs}[1]{\mathbf{#1}}
\newcommand{\reals}{\mathbb{R}}
\newcommand{\widebar}[1]{\overline{#1}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Earg}[1]{\mathbb{E}\left[{#1}\right]}
\newcommand{\Esubarg}[2]{\mathbb{E}_{#1}\left[{#2}\right]}
$$
:::

<style>
.purple { color: #7458d1ff; } /* pastel purple */
.orange { color: #fca020; } /* pastel orange */
.green { color: #3bbe67ff; } /* pastel green */
.darkblue { color: #4a9ceaff; } /* pastel dark blue */
.pink { color: #ee6ec3ff; } /* pastel pink */
</style>

```{r}
#| label: setup
#| echo: false
library(tidyverse)
library(reticulate)
theme_set(theme_classic() + theme(panel.border= element_rect(fill = NA, linewidth = .5)))
set.seed(2026)
```

```{r}
#| label: python-setup
#| echo: false
# includes the SHAP package. Can install it using,
# > conda env create -f stat479_week6.yml
# where the yaml file is located at: https://github.com/krisrs1128/stat479_notes/blob/master/notes/stat479_week6.yml
use_condaenv("stat479_week6")
```

_Readings: [1](https://arxiv.org/abs/2207.07605) (required), [2](https://proceedings.neurips.cc/paper_files/paper/2024/hash/084252ec5f05f13bf565843c1873686d-Abstract-Conference.html) (optional)_, _[Code](https://github.com/krisrs1128/stat479_notes/blob/master/notes/10-shap_computation.qmd)_

Items marked $^{\dagger}$ are not in the required reading and will not be
tested.  The required reading also covers methods beyond this handout, those
also won't be tested.

## Setup

**Goal.** Given a model $f$ and a sample $x\in\reals^{D}$, efficiently compute
the SHAP value $\varphi_d(x, f)$, the contribution of feature $d$ to the
prediction $f(x)$.

**Approach.**

There are two general strategies for efficient Shapley values computation,

   - Model-agnostic: Avoid assumptions about the model $f$. Approximate
   $\varphi_{d}(x)$ in a way that trades-off statistical accuracy for
   computation speed.
   - Model-specific: Assume a particular form for $f$ (e.g., linear or tree
   ensemble) and derive a closed-form formula that bypasses costly enumerations.
   This is often more accurate but is less general.

## Challenge

1. Let's first see why naive SHAP computation is slow. Recall the key quantities,

   * Feature $d$'s attribution,
   $$
   \varphi_d(f, x) = \sum_{S \subseteq \mathcal{D} - \{d\}} \frac{1}{D {D - 1 \choose \left|S\right|}} C(d \vert S)
   $$ {#eq-shapley}
   where $C(d \mid S) = v(S \cup \{d\}) - v(S)$ is the marginal contribution of
   $d$ to coalition $S$. Both $C$ and $v$ depend on $f$ and $x$, but we suppress this from
   our notation. We also write $\varphi_d(x)$ instead of $\varphi_d(f, x)$, since $f$ is
   fixed in these notes.

   - The value function $v(S)$ evaluates $f$ on $x$ after "removing" features
   outside of $S$ removed. We covered three removal strategies:
     - Baseline $v(S) = f(x_S, x'_{\bar{S}})$
     - Marginal^[Compared to our earlier notes, we're allowing for subsamples M < N.]: $v(S) = \mathbb{E}_{p(X_{\bar{S}})}[f(x_S, X_{\bar{S}})] \approx \frac{1}{M}\sum_{i=1}^{M} f(x_S, x_{i,\bar{S}})$
     - Conditional: $v(S) = \mathbb{E}_{p(X_{\bar{S}} \mid X_S = x_S)}[f(x_S, X_{\bar{S}})]$

1. To compute all $D$ attributions for one sample, we need $v(S)$ for every $S
\subseteq \mathcal{D}$. Using marginal feature removal, each $v(S)$ costs $M$
evaluations of $f$.  Let's try a Fermi calculation. Take $N = 100, D = 10, M =
100$. Attribution across all samples and features requires
   $$10^2 \text{ samples} \times 2^{10} \text{ subsets} \times 10^2 \text{ calls per subset} \approx 10^7 \text{ evaluations}.$$
Supposing 1ms per call, this takes $\approx 3$ hours (and this is only 100
samples with 10 features). More generally, marginal removal costs $NM\,2^D$
evaluations: exponential in $D$.

   _Exercise: How many evaluations of $f$ does $v(S)$ cost when using baseline feature removal?_

## Sampling Perspective

1. The weights in @eq-shapley define a probability distribution,
   $$
   \mathbb{P}(S) = \frac{1}{D \binom{D-1}{|S|}},
   $$
   because they are nonnegative and sum to one. This lets us rewrite
   $\varphi_d(x)$ as an expectation,
   \begin{align}
   \label{eq-distribution}
   \varphi_d(x) &= \sum_{S \subseteq \mathcal{D} \setminus \{d\}} \mathbb{P}(S)\, C(d \mid S) = \mathbb{E}_{\mathbb{P}(S)}[C(d \mid S)].
   \end{align}
   i.e., the Shapley value is the expected contribution of feature $d$ to a
   random coalition.

   Instead of enumerating all $2^{D-1}$ subsets to get the exact expectation, we
   sample $S_1, \dots, S_M \sim \mathbb{P}(S)$ and estimate
   $$
   \hat{\varphi}_d(x) = \frac{1}{M}\sum_{m=1}^{M} C(d \mid S_m).
   $$ {#eq-sampling}
   This is unbiased ($\mathbb{E}[\hat{\varphi}_d(x)] = \varphi_d(x)$) and
   ncreasing $M$ improves the approximation.

1. To draw $S_{m} \sim \mathbb{P}(S)$ in practice, we use random permutations.
Draw $\pi : \{1, \dots, D\} \to \{1, \dots, D\}$ drawn uniformly at random. For
example, for 10 features we could use,

   ```{r}
   #| label: example-perm
   pi <- sample(1:10)
   pi
   ```

   Let $\text{Pre}^{d}\left(\pi\right)$ the features appearing before $d$ in
   $\pi$. E.g., for $d = 4$,
   ```{r}
   #| label: perm-to-set
    pi[seq_len(match(4, pi) - 1)]
   ```
   It's not obvious, but $\text{Pre}^d\left(\pi\right)$ gives a sample from
   $\mathbb{P}\left(S\right)$.

1. To summarize, the procedure is,
   ```
   for m = 1, ..., M:
     sample a random permutation π_m over D features
     set S_m = Pre^d(π_m)
     compute C(d | S_m)
   return average of C(d | S_m) over m
   ```
   This is often called IME (``Interactions-based Method for Explanation'').
   Larger $M$ gives better approximations but at higher computing cost.

1. _Adaptive sampling_. The terms $C\left(d \vert S_{m}\right)$ are random
variables with mean $\varphi_{d}\left(x\right)$.

(picture of their histograms across $d$)

   For features that are irrelevant, the contributions are tightly clustered
   near zero, so even small values of $M$ are enough. Inspired by this
   observation, adaptive sampling tracks the variance of $C(d \mid S_m)$ across
   features and allocates more samples to high-variance features.

1. _Antithetic sampling_. Recall that
   $$
   \text{Var}(Z_1 + Z_2) = \text{Var}(Z_1) + \text{Var}(Z_2) + 2\text{Cov}(Z_1, Z_2).
   $$
   so negative correlated terms reduce the variance of the sum.

   _Exercise: If $\text{Cor}(Z_1, Z_2) = -1$, what is $\text{Var}(Z_1 + Z_2)$?_

   More generally, $\text{Var}\!\left(\sum_{m=1}^M Z_m\right) = \sum_m
   \text{Var}(Z_m) + \sum_{m \neq m'} \text{Cov}(Z_m, Z_{m'})$, so again
   negative pairwise correlations reduce variance.

1. Applying this to @eq-sampling, we can induce negative correlation between
$C\left(d \vert S_m\right)$ and $C\left(d \vert S_{m'}\right)$ by pairing each
$S_m$ with its complement,

   - Sample $S_m \mathbb{P}(S)$
   - Set $S_{m'} = \bar{S}_{m}$ the complement of $S$ in $\{1, \dots, D\}$.

   Despite its simplicity, this yields large variance reduction.

## Weighted Least Squares

1. A classical result in game theory [@Charnes1988] shows that SHAP values can
be characterized through a weighted least squares problem,
   $$
    \varphi\left(x\right) =
    \arg \min_{\beta \in \reals^{D + 1}} \sum_{S \subset \mathcal{D}} w(S) \left(v(S) - \beta_0 - \sum_{d \in S}\beta_{d}\right)
   $$ {#eq-kernelshap}
    where $w(S) = \left[\binom{D}{|S|} |S|(D - |S|)\right]^{-1}$ and $\varphi(x) = (\varphi_1(x), \dots, \varphi_D(x))^\top$.

1. This characterization still enumerates $2^D$ subsets, but it suggests an
approximation: sample random subsets $S_1, \dots, S_M$ and solve the weighted
regression on this reduced design. This is "KernelSHAP."

1.  Let $z_{md} = \mathbf{1}[d \in S_m]$ be the associated design matrix. The
regression suggests,
  $$
  v(S_m) - v(\emptyset) \approx \sum_d z_{md}\beta_d
  $$
  The efficiency constraint becomes,
   \begin{align*}
   \sum_d \beta_d &= v(\mathcal{D}) - v(\emptyset) \\
   &= \hat{f}(x) - \mathbb{E}[\hat{f}(X)] \\
   &=: \Delta
   \end{align*}

1. One subtlety is that $v(\emptyset) = \Esubarg{p(X)}{f(X)}$ by
definition^[Under marginal feature removal.], which is also $\beta_0$ in this
regression.  Rather than estimating this parameter freely, we fix $\hat{\beta}_0
= \frac{1}{N}f(x_i)$ and regress with this term removed.

1. Let's walk through a toy implementation of this approach. The code example at
the bottom of the notebook validates that this is equivalent to the official
implementation. The block below initializes `mask_matrix`, a matrix whose rows
will include random subsets $S_m$ stored as a binary masmk.

   ```{python}
   #| label: mask-init
   #| eval: false
   mask_matrix = np.zeros((M, D))
   ```

   We'll store our weights $W$ in `kernel_weights` and the feature-removed
   version of the sample $x$ in `synth_data`.

   ```{python}
   #| label: data-init
   #| eval: false
   kernel_weights = np.zeros(M)
   synth_data = np.tile(data, (M, 1))
   ```

1. Next, we construct the random subsets $S_{m}$ (variable `mask`) and
associated feature-removed data $(x_{S_m}, x_{\bar{S}_{m}})$.

   ```{python}
   #| label: subset-sampling
   #| eval: false
   for i in range(M):
        mask = np.random.choice([0, 1], size=D)
        mask_matrix[i] = mask
        synth_data[i * N:(i + 1) * N, mask == 1] = x[0, mask == 1]
        kernel_weights[i] = (D - np.sum(mask)) * np.sum(mask)
   ```

   The formulation @eq-kernelshap asks for the model evaluated on each of these
   feature-removed versions of the data. The variable `f_bar` is the estimate of
   $v\left(\emptyset\right)$ discussed above.
   ```{python}
   #| label: model-eval
   #| eval: false
    f_bar = np.mean(model(data))
    f = model(synth_data)
    v = np.mean(f.reshape(M, N, -1), axis=1)
   ```

1. Finally, we can run the weighted regression. But there are subtleties related
to the efficiency constraint. Notice that by the effiency constraint, $\beta_D = \Delta - \sum_{d<D} \beta_d$. This allows us to substitute into our earlier approximation,
    $$
    v(S_m) - v(\emptyset) \approx \sum_d z_{md}\beta_d \\
    $$
   to get
   $$
   v(S_m) - v(\emptyset) - z_{mD}\Delta$ \approx \sum_{d = 1}^{D - 1}z_{md}\beta_{d}
   $$
   The left hand side is `y` in the code below and the RHS is `X`.
   ```{python}
   #| label: weighted-regression
   #| eval: false
   y = (v - f_bar).flatten() - mask_matrix[:, -1] * (model(x) - f_bar)
   X = mask_matrix - mask_matrix[:, -1][:, None]
   lm = LinearRegression(fit_intercept=False).fit(X, y, sample_weight=kernel_weights)

   # final shapley values
   phi = np.zeros(D)
   phi[:-1] = lm.coef_[:-1]
   phi[-1] = (model(x) - f_bar) - np.sum(lm.coef_)
   ```

## Amortization

1. $^\dagger$ So far, we've only considered explaining a single observation $x$. In
practice, we often want to explain many observations $\{x_i\}_{i =1}^{N}$. To
speed this up, one insight is that the feature attributions across observations
are often related, and we should be able to pool across $x_i$. This will allow
us to use very rough approximations for each individual $x_i$ and still end up
with good estimates $\varphi_{d}(x_i)$ after sharing evidence.

1. $^\dagger$ This leads to the idea of _amortization_. This frames parameter estimation as
a supervised learning problem. For concreteness, imagine we've made initial estimates
   $$
   \hat{\varphi}_d^{M}(x_i) = \arg \min_{beta \in \reals^{D + 1}} \sum_{m = 1}^{M} w(S_m) \left(v(S_m) - \beta_0 - \sum_{d \in S_m}\beta_{d}\right)
   $$
   we'll assume $M$ is small, so these are noisy estimates. Then, consider the objective,
   $$
   L\left(\theta\right) = \sum_{i = 1}^{N} \left(\hat{\varphi}^{M}(x_i) - a_{\theta}(x_i)\right)^{2}
   $$
   where $a_{\theta}: \reals^{D} \to \reals$ is some model parameterized by
   $\theta$ (e.g., a linear model or neural network). This is now like a
   supervised learning problem where the "labels" are
  $\hat{\varphi}^{M}(x_i)$ and the input features are $x_i$.

1. $^\dagger$ We can apply standard optimization tools to find
  $$\hat{\theta} = \arg \min_{\theta \in \Theta} L\left(\theta\right)$$
 Once this has been found, we can very quickly get attributions for arbitrary
 test samples $x^\ast$ by just calling $a_{\hat{\theta}}(x^\ast)$. We don't even
 need to sample any subsets $S_m$ or compute any $v(S_m)$. Moreover, the
 $a_{\theta}(x^\ast)$ might be closer to the true $\varphi(x^\ast)$ than
 $\hat{\varphi}^{M}(x^\ast)$, because the regression model has learned to
 "smooth over" the noise in those initial, sample-specific estimates.

1. $^\dagger$ It's not immediately obvious that this approach is useful, because (1)
training the neural network $a_{\theta}$ can be slow and (2) if the "labels" are
too noisy, we might not learn an accurate mapping to the true $\varphi(x)$. The
research is still early, but it seems that even with moderate $M$ and simple
$a_\theta$, we might be able to learn $\varphi(x)$ with high fidelity.

## Model-Specific Versions

1. Within specific model classes, it's possible to derive formulas for
$\varphi_{d}(x_i)$ that are efficient to compute. For example, in your HW3, you
will show that
   $$
   \varphi_{d}(x) = \beta_{d}\left(x_{d} - x_{d}'\right)
   $$
   for SHAP when using baseline feature removal with baseline $x_{d}'$. The
   attribution only depends on the $d^{th}$ feature. For example, if we used the
   baseline $x' = \bar{x}$ then having a feature value $x_{d}$ that's one unit
   larger than the mean gives an attribution $\beta_{d}$. If it's one unit less
   than the mean, then feature $d$ contributes in the opposite direction, and
   the attribution becomes $-\beta_{d}$.

   A similar derivation shows that for marginal feature removal,
   $$
    \varphi_{d}(x) = \beta_{d}\left(x_{d} - \Esubarg{p(x_d)}{X_{d}}\right)
   $$
   so using the marginal approach is equivalent to using a baseline equal to the
   sample average in this particular case.

1. There are two other important cases in practice,

   - TreeSHAP: This gives SHAP values for models bsaed on trees, like CART,
   bagging, boosting, and random forest. It's implemented by the `TreeExplainer`
   class in the `shap` python package and the `treeshap` R package.
   - DeepSHAP: This derives an approximation specific to deep learning models.
   Unlke TreeSHAP, it doesn't give exact SHAP values, but it has lower variance
   than model-agnostic approaches in the deep learning case. It is implemented
   by the `DeepExplainer` class in the `shap` python package and the `DeepSHAP`
   class in the `innsight` R package.

## Code Example

1. This example shows that the `KernelSHAP` sketch above indeed agrees with the
[official
implementation](https://github.com/shap/shap/blob/61f0f8e3e0168aba1ca8f40bb8f352c37ad1519e/shap/explainers/_kernel.py#L41).
We've copied our code into the `explain` method in the class below.
`shap_values` simply loops `explain` over many samples.

   ```{python}
   #| label: kernel-explainer
   import numpy as np
   from sklearn.linear_model import LinearRegression

   class SimpleKernelExplainer:
       def __init__(self, model, data):
           self.model = model
           self.data = data
           self.N, self.D = data.shape
           self.f_bar = np.mean(model(data))

       def shap_values(self, X, M):
           explanations = []
           for i in range(X.shape[0]):
               explanations.append(self.explain(X[i:(i + 1)], M))
           return np.array(explanations)

       def explain(self, x, M):
           # initialize data structures
           mask_matrix = np.zeros((M, self.D))
           kernel_weights = np.zeros(M)
           synth_data = np.tile(self.data, (M, 1))

           # sample random subsets
           for i in range(M):
               mask = np.random.choice([0, 1], size=self.D)
               mask_matrix[i] = mask
               synth_data[i * self.N:(i + 1) * self.N, mask == 1] = x[0, mask == 1]
               kernel_weights[i] = (self.D - np.sum(mask)) * np.sum(mask)

           # model predictions on the synthetic data
           model_out = self.model(synth_data)
           v = np.mean(model_out.reshape(M, self.N, -1), axis=1)

           # weighted linear regression on adjusted f
           X = mask_matrix - mask_matrix[:, -1][:, None]
           y = (v - self.f_bar).flatten() - mask_matrix[:, -1] * (self.model(x) - self.f_bar)
           lm = LinearRegression(fit_intercept=False).fit(X, y, sample_weight=kernel_weights)

           # return shapley approximation
           phi = np.zeros(self.D)
           phi[:-1] = lm.coef_[:-1]
           phi[-1] = (self.model(x) - self.f_bar) - np.sum(lm.coef_)
           return phi
   ```

1. We'll test using a random forest model trained to fit the toy dataset
`make_friedman1` from [@Friedman1991].

   ```{python}
   #| label: fit-model
   from sklearn.ensemble import RandomForestRegressor
   from sklearn.datasets import make_friedman1

   X, y = make_friedman1(n_samples=100, n_features=5, noise=0.1)
   model = RandomForestRegressor(n_estimators=100)
   model.fit(X, y)
   ```

1. We run both our simple and the official explainer in the block below.

   ```{python}
   #| label: our-shap
   import shap

   explainer = SimpleKernelExplainer(model.predict, X)
   official_explainer = shap.KernelExplainer(model.predict, X)

   shap_values = explainer.shap_values(X, M=1000)
   official_shap_values = official_explainer.shap_values(X, nsamples=1000, silent=True)
   ```

   Their code runs a bit faster because they have extra optimizations using
   sparse rather than dense matrices to represent the mask. Nonetheless, they
   are both _much_ faster than the naive enumeration over all subsets, and their
   values roughly agree.

   ```{python}
   #| label: compare-shap

   print("Package: ", np.round(official_shap_values[:5], 4))
   print("Ours: ", np.round(shap_values[:5], 4))
   ```