---
title: "SHAP Computation"
author: Kris Sankaran
date: 2026-03-02
bibliography: references.bib
format:
  html:
    embed-resources: true
    html-math-method: mathjax
execute:
    echo: true
    message: false
    warning: false
    cache: false
---

::: {style="display: none;"}
$$
\newcommand{\bs}[1]{\mathbf{#1}}
\newcommand{\reals}{\mathbb{R}}
\newcommand{\widebar}[1]{\overline{#1}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Earg}[1]{\mathbb{E}\left[{#1}\right]}
\newcommand{\Esubarg}[2]{\mathbb{E}_{#1}\left[{#2}\right]}
$$
:::

<style>
.purple { color: #7458d1ff; } /* pastel purple */
.orange { color: #fca020; } /* pastel orange */
.green { color: #3bbe67ff; } /* pastel green */
.darkblue { color: #4a9ceaff; } /* pastel dark blue */
.pink { color: #ee6ec3ff; } /* pastel pink */
</style>

```{r}
#| label: setup
#| echo: false
library(tidyverse)
library(reticulate)
theme_set(theme_classic() + theme(panel.border= element_rect(fill = NA, linewidth = .5)))
set.seed(2026)
```

```{r}
#| label: python-setup
#| echo: false
# includes the SHAP package. Can install it using,
# > conda env create -f stat479_week6.yml
# where the yaml file is located at: https://github.com/krisrs1128/stat479_notes/blob/master/notes/stat479_week6.yml
use_condaenv("stat479_week6")
```

_Readings: [1](https://arxiv.org/abs/2207.07605) (required), [2](https://proceedings.neurips.cc/paper_files/paper/2024/hash/084252ec5f05f13bf565843c1873686d-Abstract-Conference.html) (optional)_, _[Code](https://github.com/krisrs1128/stat479_notes/blob/master/notes/10-shap_computation.qmd)_

Bullet items with $^{\dagger}$ are not in the required reading, so not tested.

## Setup

**Goal.** Given a model $f$ and a sample $x\in\reals^{D}$, efficiently compute
the SHAP value $\varphi_d(x, f)$ of the contribution of feature $d$ to the
prediction $f(x)$.

**Requirements.**

There is no single approach to speeding up computation of Shapley values in
machine learning. We'll survey five of the methods discussed in the reading and
list the requirements for that particular method once we reach it.

## Challenge

1. Before we try speeding up SHAP computation, let's see how bad a naive
approach can be. First, let's review the key quantities in applying SHAP for
machine learning.

   * Feature $d$'s attribution,
   $$
   \varphi_d(f, x) = \sum_{S \subseteq \mathcal{D} - \{d\}} \frac{1}{D {D - 1 \choose \left|S\right|}} C(d \vert S)
   $$ {#eq-shapley}
   where $C\left(d \vert S\right)$ is the contribution of $d$ to coalition $S$,
   $$C\left(d \vert S\right) = v(S \cup \{d\}) - v(S).$$
   This depends on both $f$ and $x$, but we're suppressing it from the
    notation.

   * The functions $v$ were defined by calling $f$ on a version of $x$ where the features in coordinates $\bar{S}$ had been removed. There were three approaches,

        - Baseline removal: $v(S) = f(x_{S}, x'_{\bar{S}})$
        - Marginal removal^[We're allowing for subsamples M < N in our approximation.]: $v(S) = \Esubarg{p(X_{\bar{S}})}{f(x_{S}, X_{\bar{S}})} \approx \frac{1}{M}\sum_{i = 1}^{M} f\left(x_S, x_{i,\bar{S}}\right)$.
        - Conditional removal^[We'll ignore this approach for the rest of the lecture, since there is no universal approach to approximating this conditional expectation.]: $v(S) = \Esubarg{p(X_{\bar{S}} \vert X_{S} = x_{S})}{f(x_{S}, X_{\bar{S}})}$

1. To compute all $D$ attributions for one sample, we need $v(S)$ for every $S
\subseteq \mathcal{D}$, and each $v(S)$ costs $M$ evaluations of $f$, supposing
marginal feature removal. Let's try a Fermi calculation. Consider $N = 100, D =
10, M = 100$. To get attributions across all $N$ samples and $D$ features, we'll
need $10 ^ 2 \text{ samples} \times 10 ^ 2 \text{ calls} \times 2^{D} \text{
subsets} \approx 10 ^ 7$ function calls. If each call takes 1ms, then this would
take $\approx 3$ hours. This is only 100 samples and 10 features. More
generally, marginal feature removal requires $N M 2^D$ evaluations, so the cost
grows exponentially in $D$.

   _Exercise: How many evaluations of $f$ does $v(S)$ cost when using baseline feature removal?_

## Sampling Subsets

**Requirements.**

## Weighted Least Squares

**Requirements.**

## Local Shapley

**Requirements.**

## Stochastic Amortization

**Requirements.**

## Model-Specific Versions

**Requirements.**