---
title: "SHAP Computation"
author: Kris Sankaran
date: 2026-03-02
bibliography: references.bib
format:
  html:
    embed-resources: true
    html-math-method: mathjax
execute:
    echo: true
    message: false
    warning: false
    cache: false
---

::: {style="display: none;"}
$$
\newcommand{\bs}[1]{\mathbf{#1}}
\newcommand{\reals}{\mathbb{R}}
\newcommand{\widebar}[1]{\overline{#1}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Earg}[1]{\mathbb{E}\left[{#1}\right]}
\newcommand{\Esubarg}[2]{\mathbb{E}_{#1}\left[{#2}\right]}
$$
:::

<style>
.purple { color: #7458d1ff; } /* pastel purple */
.orange { color: #fca020; } /* pastel orange */
.green { color: #3bbe67ff; } /* pastel green */
.darkblue { color: #4a9ceaff; } /* pastel dark blue */
.pink { color: #ee6ec3ff; } /* pastel pink */
</style>

```{r}
#| label: setup
#| echo: false
library(tidyverse)
library(reticulate)
theme_set(theme_classic() + theme(panel.border= element_rect(fill = NA, linewidth = .5)))
set.seed(2026)
```

```{r}
#| label: python-setup
#| echo: false
# includes the SHAP package. Can install it using,
# > conda env create -f stat479_week6.yml
# where the yaml file is located at: https://github.com/krisrs1128/stat479_notes/blob/master/notes/stat479_week6.yml
use_condaenv("stat479_week6")
```

_Readings: [1](https://arxiv.org/abs/2207.07605) (required), [2](https://proceedings.neurips.cc/paper_files/paper/2024/hash/084252ec5f05f13bf565843c1873686d-Abstract-Conference.html) (optional)_, _[Code](https://github.com/krisrs1128/stat479_notes/blob/master/notes/10-shap_computation.qmd)_

Bullet items with $^{\dagger}$ are not in the required reading, so not tested.

## Setup

**Goal.** Given a model $f$ and a sample $x\in\reals^{D}$, efficiently compute
the SHAP value $\varphi_d(x, f)$ of the contribution of feature $d$ to the
prediction $f(x)$.

**Approach.**

At a high-level, there are two approaches to efficiently calculating Shapley
attributions.

   - Model-agnostic: Without making any assumptions about the model $f$, find a way of approximating $\varphi_{d}(x)$ that trade-offs statistical accuracy with computational efficiency.
    - Model-specific: Assuming a particular form for $f$ (e.g., linearity or
    tree ensemble), derive a model-specific formula for Shapley attributions
    that bypasses the enumeration over subsets required in the general case.

The basic trade-off is,
   - Model-agnostic: Works on any model, but only approximates $\varphi_{d}(x)$
   - Model-specific: Works only on narrow model classes, but gives exact $\varphi_{d}(x)$.

## Challenge

1. Before we try speeding up SHAP computation, let's see why it's slow.  Recall
the main quantities of interest,

   * Feature $d$'s attribution,
   $$
   \varphi_d(f, x) = \sum_{S \subseteq \mathcal{D} - \{d\}} \frac{1}{D {D - 1 \choose \left|S\right|}} C(d \vert S)
   $$ {#eq-shapley}
   where $C\left(d \vert S\right)$ is the contribution of $d$ to coalition $S$,
   $$C\left(d \vert S\right) = v(S \cup \{d\}) - v(S).$$
   This depends on both $f$ and $x$, but we're suppressing it from the
    notation.

   * The functions $v$ were defined by calling $f$ on a version of $x$ where the features in coordinates $\bar{S}$ had been removed. There were three approaches,

        - Baseline removal: $v(S) = f(x_{S}, x'_{\bar{S}})$
        - Marginal removal^[We're allowing for subsamples M < N in our approximation.]: $v(S) = \Esubarg{p(X_{\bar{S}})}{f(x_{S}, X_{\bar{S}})} \approx \frac{1}{M}\sum_{i = 1}^{M} f\left(x_S, x_{i,\bar{S}}\right)$.
        - Conditional removal^[We'll ignore this approach for the rest of the lecture, since there is no universal approach to approximating this conditional expectation.]: $v(S) = \Esubarg{p(X_{\bar{S}} \vert X_{S} = x_{S})}{f(x_{S}, X_{\bar{S}})}$

1. To compute all $D$ attributions for one sample, we need $v(S)$ for every $S
\subseteq \mathcal{D}$, and each $v(S)$ costs $M$ evaluations of $f$, supposing
marginal feature removal. Let's try a Fermi calculation. Consider $N = 100, D =
10, M = 100$. To get attributions across all $N$ samples and $D$ features, we'll
need $10 ^ 2 \text{ samples} \times 10 ^ 2 \text{ calls} \times 2^{D} \text{
subsets} \approx 10 ^ 7$ function calls. If each call takes 1ms, then this would
take $\approx 3$ hours. This is only 100 samples and 10 features. More
generally, marginal feature removal requires $N M 2^D$ evaluations, so the cost
grows exponentially in $D$.

   _Exercise: How many evaluations of $f$ does $v(S)$ cost when using baseline feature removal?_

## Sampling Subsets

1. We can interpret the weights $\frac{1}{D {D - 1 \choose \left|S\right|}}$ as
probabilities, since they are nonnegative and sum to one. From this point of
view,
   $$
   \varphi_d(f, x) = \sum_{S \subseteq \mathcal{D} - \{d\}} \mathbb{P}\left(S\right) C(d \vert S) \\
   &= \Esubarg{\mathbb{P}(S)}{C(d \vert S)} \\
   &= \Esubarg{\mathbb{P}{S}}{\text{Contribution of feature } d \text{ to coalition} S}.
   $$ {#eq-distribution}
   for a distribution $\mathbb{P}$ over subsets. Instead of enumerating over all
   subsets to get the exact expectation, we could sample random subsets
   according to $S_{m} \sim \pi(S)$ and then estimate,
   $$
   \hat{\varphi}_{d}(f, x) = \sum_{m = 1}^{M} C(d \vert S_{m})
   $$ {#eq-sampling}
   By the identity above, this is an unbiased estimator,
   $$\Esubarg{\mathbb{P}(S)}{\hat{\varphi}_{d}(f, x)} = \varphi_{d}(f, x)$$
   and the larger the $M$, the better an approximation we obtain.

1. To draw $S_{m} \sim \mathbb{P}(S)$ in practice, we usually use permutations.
Specifically, consider a permutation over features $\pi : \{1, \dots, D\} \to
\{1, \dots, D\}$ drawn uniformly at random. For example, if we had 10 features,
we could use,

   ```{r}
   pi <- sample(1:10)
   pi
   ```

   Let $\text{Pre}^{d}\left(\pi\right)$ be the items in the permutation up until
   the $d^th$ feature shows up. For example, for $d = 4$, we would consider,
   ```{r}
    pi[seq_len(match(4, pi) - 1)]
   ```
   It's not obvious, but the set $\text{Pre}^d\left(\pi\right)$ gives a draw
   from $\mathbb{P}\left(S\right)$.

1. To summarize, the procedure is,

   ```
   for m = 1, ..., M:
     sample a random permutation π_m over D features
     set S_m = Pre^d(π_m)
     compute C(d | S_m)
   return average of C(d | S_m) over m
   ```

   This is often called the Interactions-base

1. Sampling more permutations $M$ leads to a better approximation, at the cost
of higher computation.

1. _Adaptive sampling_. The terms $C\left(d \vert S_{m}\right)$ are random
variables with mean $\varphi_{d}(f, x\right)$.  (picture of their histograms
across $d$) For some choices of $d$, $C\left(d \vert S_{m}\right)$ may have low
variance (this is especially true if a feature isn't relevant to the prediction
and all the contributions are near zero).  In this case, even averaging a small
number of $M$ yields a good approximation of $\varphi_{d}(f, x)$.  Adaptive
sampling keeps track of the variance in $C(d \vert S_{m}\right)$ across $d$ and
allocates more samples to the features with higher variance.

1. _Antithetic sampling_. Remember from probability, that the variance of a sum of two variables is,
   $$
   \text{Var}(Z_1, Z_2) = \text{Var}(Z_1) + \text{Var}(Z_2) + \text{Cov}(Z_1, Z_2).
   $$
   The implication is that we can reduce the variance of a sum by choosing the
   variables so that they are negatively correlated.

   _Exercise: If $\text{Cor}(Z_1, Z_2) = -1$, what is the variance of the sum?_

   This holds for larger sums too. For general $M$,
   $$
    \text{Var}\left(\sum_{m = 1}^{M} Z_{m}\right) = \sum_{m = 1}^{M}\text{Var}(Z_m) + \sum_{m \neq m'} \text{Cov}(Z_m, Z_{m'})
   $$
   and again we can reduce the variance of the sum by choosing the $Z_m$ with negative pairwise correlations.

1. This approach can be applied to @eq-sampling. To create negative correlation
between the terms $C(d \vert S_m)$ and $C(d \vert S_{m'}\right)$, we pairs each
$S_m$ with its complement,

   - Sample $S_m \mathbb{P}(S)$
   - Set $S_{m'} &= \bar{S}_{m}$ the complement of $S$ among the features $\{1, \dots, D\}$.

   Though this is simple, it yields competitive performance in practice.

## Weighted Least Squares

1. So far, we've relied on the representation @eq-distribution. There is another
representation that also suggests a natural approximation to the SHAP value. A
classic result in game theory [@Charnes1988] shows that SHAP values can be
expressed as the solution to a weighted least squares problem.
   $$
    \begin{bmatrix} \varphi_{1}\left(f, x\right) \\ \vdots \\ \varphi_{D}\left(f, x\right) \end{bmatrix}&=
    \arg \min_{beta \in \reals^{D + 1}} \sum_{S \subset \mathcal{D}} w(S) \left(v(S) - \beta_0 - \sum_{d \in S}\beta_{d})
   $$
   where the regression weights are given by $W(S) = (D - 1) / ({D \choose \left|S\right|})(|S|(|D| - |S|))$.

1. For example, suppose that we had only two features. The regression has the
form,

   More generally,

1. This theoretical result relies on all subsets $S \subset \mathcal{D}$, but it
suggests an approximation: Sample random subsets $S$ and run the regression onto
the reduced design matrix. This approximate version of weighted lienar regression is called "KernelSHAP."

1. Let's walk through a toy implementation of this approach. First we initialize
`mask_matrix`, a matrix whose rows will include random subsets $S_m$ stored as a
binary masmk.

   ```{r, eval=FALSE}
   # initialize data structures
   mask_matrix <- matrix(0, nrow = nsamples, ncol = M)
   ```

   We'll store our weights $W$ in `kernel_weights` and the feature-removed
   version of the sample $x$ in `synth_data`.

   ```{r}
   kernel_weights <- numeric(nsamples)
   synth_data <- data[rep(seq_len(N), nsamples), ]
   ```

   ```{r}
   # keep values at random subsets of features
   for (i in seq_len(nsamples)) {
     mask <- sample(0:1, M, replace = TRUE)
     mask_matrix[i, ] <- mask
     rows <- seq((i - 1) * N + 1, i * N)
     synth_data[rows, mask == 1] <- x[1, mask == 1]
     kernel_weights[i] <- (M - sum(mask)) * sum(mask)
   }

   # model predictions on masked data
   model_out <- model(synth_data)
   f <- rowMeans(matrix(model_out, nrow = nsamples, ncol = N, byrow = TRUE)) - f_bar

   # weighted regression on adjusted f
   last_col <- mask_matrix[, M]
   X <- mask_matrix - last_col
   y <- f - last_col * (model(x) - f_bar)
   lm_fit <- lm.wfit(x = X, y = y, w = kernel_weights)
   ```


## Local Shapley


## Stochastic Amortization


## Model-Specific Versions
