---
title: "SHAP Computation"
author: Kris Sankaran
date: 2026-03-02
bibliography: references.bib
format:
  html:
    embed-resources: true
    html-math-method: mathjax
execute:
    echo: true
    message: false
    warning: false
    cache: false
    eval: false
---

::: {style="display: none;"}
$$
\newcommand{\bs}[1]{\mathbf{#1}}
\newcommand{\reals}{\mathbb{R}}
\newcommand{\widebar}[1]{\overline{#1}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Earg}[1]{\mathbb{E}\left[{#1}\right]}
\newcommand{\Esubarg}[2]{\mathbb{E}_{#1}\left[{#2}\right]}
$$
:::

<style>
.purple { color: #7458d1ff; } /* pastel purple */
.orange { color: #fca020; } /* pastel orange */
.green { color: #3bbe67ff; } /* pastel green */
.darkblue { color: #4a9ceaff; } /* pastel dark blue */
.pink { color: #ee6ec3ff; } /* pastel pink */
</style>

```{r}
#| label: setup
#| echo: false
library(tidyverse)
library(reticulate)
theme_set(theme_classic() + theme(panel.border= element_rect(fill = NA, linewidth = .5)))
set.seed(2026)
```

```{r}
#| label: python-setup
#| echo: false
# includes the SHAP package. Can install it using,
# > conda env create -f stat479_week6.yml
# where the yaml file is located at: https://github.com/krisrs1128/stat479_notes/blob/master/notes/stat479_week6.yml
use_condaenv("stat479_week6")
```

_Readings: [1](https://arxiv.org/abs/2207.07605) (required), [2](https://proceedings.neurips.cc/paper_files/paper/2024/hash/084252ec5f05f13bf565843c1873686d-Abstract-Conference.html) (optional)_, _[Code](https://github.com/krisrs1128/stat479_notes/blob/master/notes/10-shap_computation.qmd)_

Items marked $^{\dagger}$ are not in the required reading and will not be
tested.  The required reading also covers methods beyond this handout, those
also won't be tested.

## Setup

**Goal.** Given a model $f$ and a sample $x\in\reals^{D}$, efficiently compute
the SHAP value $\varphi_d(x, f)$, the contribution of feature $d$ to the
prediction $f(x)$.

**Approach.**

There are two general strategies for efficient Shapley values computation,

   - Model-agnostic: Without making any assumptions about the model $f$, find a
   way of approximating $\varphi_{d}(x)$ that trade-offs statistical accuracy
   with computational efficiency.
   - Model-specific: Assuming a particular form for $f$ (e.g., linearity or tree
   ensemble), derive a model-specific formula for Shapley attributions that
   bypasses the enumeration over subsets required in the general case. This can
   often be much more accurate, but of course makes stronger assumptions.

## Challenge

1. Before we try speeding up SHAP computation, let's see why it's slow.  Recall
the main quantities of interest,

   * Feature $d$'s attribution,
   $$
   \varphi_d(f, x) = \sum_{S \subseteq \mathcal{D} - \{d\}} \frac{1}{D {D - 1 \choose \left|S\right|}} C(d \vert S)
   $$ {#eq-shapley}
   where $C\left(d \vert S\right)$ is the contribution of $d$ to coalition $S$,
   $$C\left(d \vert S\right) = v(S \cup \{d\}) - v(S).$$
   This depends on both $f$ and $x$, but we're suppressing it from the notation.
   We'll also write $\varphi_{d}(x)$ instead of $\varphi_{d}(f, x)$ since the
   function to explain is fixed throughout these notes.

   * The functions $v$ were defined by calling $f$ on a version of $x$ where the features in coordinates $\bar{S}$ had been removed. There were three approaches,

        - Baseline removal: $v(S) = f(x_{S}, x'_{\bar{S}})$
        - Marginal removal^[We're allowing for subsamples M < N in our approximation.]: $v(S) = \Esubarg{p(X_{\bar{S}})}{f(x_{S}, X_{\bar{S}})} \approx \frac{1}{M}\sum_{i = 1}^{M} f\left(x_S, x_{i,\bar{S}}\right)$.
        - Conditional removal^[We'll ignore this approach for the rest of the lecture, since there is no universal approach to approximating this conditional expectation.]: $v(S) = \Esubarg{p(X_{\bar{S}} \vert X_{S} = x_{S})}{f(x_{S}, X_{\bar{S}})}$

1. To compute all $D$ attributions for one sample, we need $v(S)$ for every $S
\subseteq \mathcal{D}$, and each $v(S)$ costs $M$ evaluations of $f$, supposing
marginal feature removal. Let's try a Fermi calculation. Consider $N = 100, D =
10, M = 100$. To get attributions across all $N$ samples and $D$ features, we'll
need $10 ^ 2 \text{ samples} \times 10 ^ 2 \text{ calls} \times 2^{D} \text{
subsets} \approx 10 ^ 7$ function calls. If each call takes 1ms, then this would
take $\approx 3$ hours. This is only 100 samples and 10 features. More
generally, marginal feature removal requires $N M 2^D$ evaluations, so the cost
grows exponentially in $D$.

   _Exercise: How many evaluations of $f$ does $v(S)$ cost when using baseline feature removal?_

## Sampling Subsets

1. We can interpret the weights in @eq-shapley as weights,
   $$
   \mathbb{P}\left(S\right) = 1/({D {D - 1 \choose \left|S\right|}})
   $$
  since they are nonnegative and sum to one. From this point of view,
   \begin{align}
   \label{eq-distribution}
   \varphi_d(x) &= \sum_{S \subseteq \mathcal{D} - \{d\}} \mathbb{P}\left(S\right) C(d \vert S) \\
   &= \Esubarg{\mathbb{P}(S)}{C(d \vert S)} \\
   &= \text{``Expected contribution of feature } d \text{ to a random coalition''}.
   \end{align}
   Instead of enumerating over all subsets to get the exact expectation, we
   could sample random subsets according to $S_{m} \sim \mathbb{P}(S)$ and then
   estimate,
   $$
   \hat{\varphi}_{d}(x) = \sum_{m = 1}^{M} C(d \vert S_{m})
   $$ {#eq-sampling}
   By the derivation in @eq-distribution, this is an unbiased estimator, meaning,
   $$\Esubarg{\mathbb{P}(S)}{\hat{\varphi}_{d}(x)} = \varphi_{d}(x)$$
   and the larger the $M$, the better an approximation we obtain.

1. To draw $S_{m} \sim \mathbb{P}(S)$ in practice, we usually use permutations.
Specifically, consider a permutation over features $\pi : \{1, \dots, D\} \to
\{1, \dots, D\}$ drawn uniformly at random. For example, if we had 10 features,
we could use,

   ```{r}
   #| label: example-perm
   pi <- sample(1:10)
   pi
   ```

   Let $\text{Pre}^{d}\left(\pi\right)$ be the items in the permutation up until
   the $d^{th}$ feature shows up. For example, for $d = 4$, we would consider,
   ```{r}
   #| label: perm-to-set
    pi[seq_len(match(4, pi) - 1)]
   ```
   It's not obvious, but the set $\text{Pre}^d\left(\pi\right)$ gives a draw
   from $\mathbb{P}\left(S\right)$.

1. To summarize, the procedure is,

   ```
   for m = 1, ..., M:
     sample a random permutation π_m over D features
     set S_m = Pre^d(π_m)
     compute C(d | S_m)
   return average of C(d | S_m) over m
   ```

   This is often called the Interactions-base

1. Sampling more permutations $M$ leads to a better approximation, at the cost
of higher computation.

1. _Adaptive sampling_. The terms $C\left(d \vert S_{m}\right)$ are random
variables with mean $\varphi_{d}\left(x\right)$.  (picture of their histograms
across $d$) For some choices of $d$, $C\left(d \vert S_{m}\right)$ may have low
variance (this is especially true if a feature isn't relevant to the prediction
and all the contributions are near zero).  In this case, even averaging a small
number of $M$ yields a good approximation of $\varphi_{d}(x)$.  Adaptive
sampling keeps track of the variance in $C\left(d \vert S_{m}\right)$ across $d$ and
allocates more samples to the features with higher variance.

1. _Antithetic sampling_. Remember from probability, that the variance of a sum of two variables is,
   $$
   \text{Var}(Z_1, Z_2) = \text{Var}(Z_1) + \text{Var}(Z_2) + \text{Cov}(Z_1, Z_2).
   $$
   The implication is that we can reduce the variance of a sum by choosing the
   variables so that they are negatively correlated.

   _Exercise: If $\text{Cor}(Z_1, Z_2) = -1$, what is the variance of the sum?_

   This holds for larger sums too. For general $M$,
   $$
    \text{Var}\left(\sum_{m = 1}^{M} Z_{m}\right) = \sum_{m = 1}^{M}\text{Var}(Z_m) + \sum_{m \neq m'} \text{Cov}(Z_m, Z_{m'})
   $$
   and again we can reduce the variance of the sum by choosing the $Z_m$ with negative pairwise correlations.

1. This approach can be applied to @eq-sampling. To create negative correlation
between the terms $C\left(d \vert S_m\right)$ and $C\left(d \vert S_{m'}\right)$, we pairs each
$S_m$ with its complement,

   - Sample $S_m \mathbb{P}(S)$
   - Set $S_{m'} = \bar{S}_{m}$ the complement of $S$ among the features $\{1, \dots, D\}$.

   Though this is simple, it yields large improvements.

## Weighted Least Squares

1. So far, we've relied on the representation @eq-distribution. There is another
representation that also suggests a natural approximation to the SHAP value. A
classic result in game theory [@Charnes1988] shows that SHAP values can be
expressed as the solution to a weighted least squares problem.
   $$
    \varphi\left(x\right) =
    \arg \min_{\beta \in \reals^{D + 1}} \sum_{S \subset \mathcal{D}} w(S) \left(v(S) - \beta_0 - \sum_{d \in S}\beta_{d}\right)
   $$ {#eq-kernelshap}
   where the regression weights are given by $w(S) = \left[{D \choose
   \left|S\right|} \left|S\right|\left(D - \left|S\right|\right)\right]^{-1}$
   and we're concatenating the SHAP values into,
   $$
   \varphi(x) = \begin{bmatrix} \varphi_{1}\left(x\right) \\ \vdots \\ \varphi_{D}\left(x\right) \end{bmatrix}
   $$

1. This theoretical result relies on all subsets $S \subset \mathcal{D}$, but it
suggests an approximation: Sample random subsets $S_m$ and run the regression
onto the reduced design matrix. This approximate version of weighted linear
regression is called "KernelSHAP."

1. Let's walk through a toy implementation of this approach. The code example at
the bottom of the notebook validates that this is equivalent to the official
implementation. The block below initializes `mask_matrix`, a matrix whose rows
will include random subsets $S_m$ stored as a binary masmk.

   ```{python}
   #| label: mask-init
   #| eval: false
   mask_matrix = np.zeros((M, D))
   ```

   We'll store our weights $W$ in `kernel_weights` and the feature-removed
   version of the sample $x$ in `synth_data`.

   ```{python}
   #| label: data-init
   #| eval: false
   kernel_weights = np.zeros(M)
   synth_data = np.tile(data, (M, 1))
   ```

1. Next, we construct the random subsets $S_{m}$ (variable `mask`) and
associated feature-removed data $(x_{S_m}, x_{\bar{S}_{m}})$.

   ```{python}
   #| label: subset-sampling
   #| eval: false
   for i in range(M):
        mask = np.random.choice([0, 1], size=D)
        mask_matrix[i] = mask
        synth_data[i * N:(i + 1) * N, mask == 1] = x[0, mask == 1]
        kernel_weights[i] = (D - np.sum(mask)) * np.sum(mask)
   ```

   The formulation @eq-kernelshap asks for the model evaluated on each of these
   feature-removed versions of the data.
   ```{python}
   #| label: model-eval
   #| eval: false
    f_bar = np.mean(model(data))
    f = model(synth_data)
    v = np.mean(f.reshape(M, N, -1), axis=1)
   ```

1. Finally, we can run the weighted regression. By definition,
$\varphi_{\emptyset}$ is the mean of $\hat{f}(x_i)$ over $i$, so we fix that
effect before running the regression.

   ```{python}
   #| label: weighted-regression
   #| eval: false
   X = mask_matrix - mask_matrix[:, -1][:, None]
   y = (v - f_bar).flatten() - mask_matrix[:, -1] * (model(x) - f_bar)
   lm = LinearRegression(fit_intercept=False).fit(X, y, sample_weight=kernel_weights)

   # final shapley values
   phi = np.zeros(D)
   phi[:-1] = lm.coef_[:-1]
   phi[-1] = (model(x) - f_bar) - np.sum(lm.coef_)
   ```

## Amortization

1. $^\dagger$ So far, we've only considered explaining a single observation $x$. In
practice, we often want to explain many observations $\{x_i\}_{i =1}^{N}$. To
speed this up, one insight is that the feature attributions across observations
are often related, and we should be able to pool across $x_i$. This will allow
us to use very rough approximations for each individual $x_i$ and still end up
with good estimates $\varphi_{d}(x_i)$ after sharing evidence.

1. $^\dagger$ This leads to the idea of _amortization_. This frames parameter estimation as
a supervised learning problem. For concreteness, imagine we've made initial estimates
   $$
   \hat{\varphi}_d^{M}(x_i) = \arg \min_{beta \in \reals^{D + 1}} \sum_{m = 1}^{M} w(S_m) \left(v(S_m) - \beta_0 - \sum_{d \in S_m}\beta_{d}\right)
   $$
   we'll assume $M$ is small, so these are noisy estimates. Then, consider the objective,
   $$
   L\left(\theta\right) = \sum_{i = 1}^{N} \left(\hat{\varphi}^{M}(x_i) - a_{\theta}(x_i)\right)^{2}
   $$
   where $a_{\theta}: \reals^{D} \to \reals$ is some model parameterized by
   $\theta$ (e.g., a linear model or neural network). This is now like a
   supervised learning problem where the "labels" are
  $\hat{\varphi}^{M}(x_i)$ and the input features are $x_i$.

1. $^\dagger$ We can apply standard optimization tools to find
  $$\hat{\theta} = \arg \min_{\theta \in \Theta} L\left(\theta\right)$$
 Once this has been found, we can very quickly get attributions for arbitrary
 test samples $x^\ast$ by just calling $a_{\hat{\theta}}(x^\ast)$. We don't even
 need to sample any subsets $S_m$ or compute any $v(S_m)$. Moreover, the
 $a_{\theta}(x^\ast)$ might be closer to the true $\varphi(x^\ast)$ than
 $\hat{\varphi}^{M}(x^\ast)$, because the regression model has learned to
 "smooth over" the noise in those initial, sample-specific estimates.

1. $^\dagger$ It's not immediately obvious that this approach is useful, because (1)
training the neural network $a_{\theta}$ can be slow and (2) if the "labels" are
too noisy, we might not learn an accurate mapping to the true $\varphi(x)$. The
research is still early, but it seems that even with moderate $M$ and simple
$a_\theta$, we might be able to learn $\varphi(x)$ with high fidelity.

## Model-Specific Versions

1. Within specific model classes, it's possible to derive formulas for
$\varphi_{d}(x_i)$ that are efficient to compute. For example, in your HW3, you
will show that
   $$
   \varphi_{d}(x) = \beta_{d}\left(x_{d} - x_{d}'\right)
   $$
   for SHAP when using baseline feature removal with baseline $x_{d}'$. The
   attribution only depends on the $d^{th}$ feature. For example, if we used the
   baseline $x' = \bar{x}$ then having a feature value $x_{d}$ that's one unit
   larger than the mean gives an attribution $\beta_{d}$. If it's one unit less
   than the mean, then feature $d$ contributes in the opposite direction, and
   the attribution becomes $-\beta_{d}$.

   A similar derivation shows that for marginal feature removal,
   $$
    \varphi_{d}(x) = \beta_{d}\left(x_{d} - \Esubarg{p(x_d)}{X_{d}}\right)
   $$
   so using the marginal approach is equivalent to using a baseline equal to the
   sample average in this particular case.

1. There are two other important cases in practice,

   - TreeSHAP: This gives SHAP values for models bsaed on trees, like CART,
   bagging, boosting, and random forest. It's implemented by the `TreeExplainer`
   class in the `shap` python package and the `treeshap` R package.
   - DeepSHAP: This derives an approximation specific to deep learning models.
   Unlke TreeSHAP, it doesn't give exact SHAP values, but it has lower variance
   than model-agnostic approaches in the deep learning case. It is implemented
   by the `DeepExplainer` class in the `shap` python package and the `DeepSHAP`
   class in the `innsight` R package.

## Code Example

1. This example shows that the `KernelSHAP` sketch above indeed agrees with the
[official
implementation](https://github.com/shap/shap/blob/61f0f8e3e0168aba1ca8f40bb8f352c37ad1519e/shap/explainers/_kernel.py#L41).
We've copied our code into the `explain` method in the class below.
`shap_values` simply loops `explain` over many samples.

   ```{python}
   #| label: kernel-explainer
   import numpy as np
   from sklearn.linear_model import LinearRegression

   class SimpleKernelExplainer:
       def __init__(self, model, data):
           self.model = model
           self.data = data
           self.N, self.D = data.shape
           self.f_bar = np.mean(model(data))

       def shap_values(self, X, M):
           explanations = []
           for i in range(X.shape[0]):
               explanations.append(self.explain(X[i:(i + 1)], M))
           return np.array(explanations)

       def explain(self, x, M):
           # initialize data structures
           mask_matrix = np.zeros((M, self.D))
           kernel_weights = np.zeros(M)
           synth_data = np.tile(self.data, (M, 1))

           # sample random subsets
           for i in range(M):
               mask = np.random.choice([0, 1], size=self.D)
               mask_matrix[i] = mask
               synth_data[i * self.N:(i + 1) * self.N, mask == 1] = x[0, mask == 1]
               kernel_weights[i] = (self.D - np.sum(mask)) * np.sum(mask)

           # model predictions on the synthetic data
           model_out = self.model(synth_data)
           v = np.mean(model_out.reshape(M, self.N, -1), axis=1)

           # weighted linear regression on adjusted f
           X = mask_matrix - mask_matrix[:, -1][:, None]
           y = (v - self.f_bar).flatten() - mask_matrix[:, -1] * (self.model(x) - self.f_bar)
           lm = LinearRegression(fit_intercept=False).fit(X, y, sample_weight=kernel_weights)

           # return shapley approximation
           phi = np.zeros(self.D)
           phi[:-1] = lm.coef_[:-1]
           phi[-1] = (self.model(x) - self.f_bar) - np.sum(lm.coef_)
           return phi
   ```

1. We'll test using a random forest model trained to fit the toy dataset
`make_friedman1` from [@Friedman1991].

   ```{python}
   #| label: fit-model
   from sklearn.ensemble import RandomForestRegressor
   from sklearn.datasets import make_friedman1

   X, y = make_friedman1(n_samples=100, n_features=5, noise=0.1)
   model = RandomForestRegressor(n_estimators=100)
   model.fit(X, y)
   ```

1. We run both our simple and the official explainer in the block below.

   ```{python}
   #| label: our-shap
   import shap

   explainer = SimpleKernelExplainer(model.predict, X)
   official_explainer = shap.KernelExplainer(model.predict, X)

   shap_values = explainer.shap_values(X, M=1000)
   official_shap_values = official_explainer.shap_values(X, nsamples=1000, silent=True)
   ```

   Their code runs a bit faster because they have extra optimizations using
   sparse rather than dense matrices to represent the mask. Nonetheless, they
   are both _much_ faster than the naive enumeration over all subsets, and their
   values roughly agree.

   ```{python}
   #| label: compare-shap

   print("Package: ", np.round(official_shap_values[:5], 4))
   print("Ours: ", np.round(shap_values[:5], 4))
   ```