---
title: "Introduction to Deep Learning"
author: Kris Sankaran
date: 2026-03-16
bibliography: references.bib
format:
  html:
    embed-resources: true
    html-math-method: mathjax
    fig-align: center
execute:
    echo: true
    message: false
    warning: false
    cache: false
    eval: true
---

::: {style="display: none;"}
$$
\newcommand{\bs}[1]{\mathbf{#1}}
\newcommand{\reals}{\mathbb{R}}
\newcommand{\widebar}[1]{\overline{#1}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Earg}[1]{\mathbb{E}\left[{#1}\right]}
\newcommand{\Esubarg}[2]{\mathbb{E}_{#1}\left[{#2}\right]}
$$
:::

<style>
.purple { color: #7458d1ff; } /* pastel purple */
.orange { color: #fca020; } /* pastel orange */
.green { color: #3bbe67ff; } /* pastel green */
.darkblue { color: #4a9ceaff; } /* pastel dark blue */
.pink { color: #ee6ec3ff; } /* pastel pink */
</style>

```{r}
#| label: setup
#| echo: false
library(tidyverse)
library(reticulate)
theme_set(theme_classic() + theme(panel.border= element_rect(fill = NA, linewidth = .5)))
set.seed(2026)
```

```{r}
#| label: python-setup
#| echo: false
# includes the SHAP package. Can install it using,
# > conda env create -f stat479_week6.yml
# where the yaml file is located at: https://github.com/krisrs1128/stat479_notes/blob/master/notes/stat479_week9.yml
#use_condaenv("stat479_week9")
```

_Readings: [1](https://visionbook.mit.edu/neural_nets.html) (sections 12.1 - 12.7, but skip 12.5.2 - 12.5.3), [2](https://docs.pytorch.org/tutorials/beginner/basics/intro.html#learn-the-basics) (optional)_, _[Code](https://github.com/krisrs1128/stat479_notes/blob/master/notes/11-deep_learning_overview.qmd)_

Items marked $^{\dagger}$ are not in the required reading and will not be
tested.  The required reading covers some methods beyond this handout, those
also won't be tested.

## Setup

**Goal.** Given data $\{\left({x_i, y_i}\right)\}_{i = 1}^{N}$, develop a predictor $f$ for $y_{i}$.

  - More importantly, create a generic recipe for learning predictors $f$ that works as soon as we have enough examples $x_i$ and $y_i$, with as little manual effort as possible.

**Requirements.**

- The recipe should apply to a variety of data structures out-of-the-box. See the table below.
- The method must handle highly nonlinear relationships.
- The method must scale to very large $N$.

| Task | $x_i$ | $y_i$ | Example |
|---|---|---| --- | --- |
| Image Classification | image | label | |
| Object Detection | image | label and bounding box | |
| Language Modeling | sequence of words | next word | |
| Sentence Translation | sequence of words | sequence of words | |
| Style Transfer | image | new image | |


**Approach.** We represent data as generic $D$-dimensional tensors and define parameterized modules $f_l(\cdot; \theta_l)$ that process them. Composing these modules creates architectures whose intermediate representations $h_l$ link the inputs $x$ with outputs $y$. Parameters $\theta_l$ are fit by optimizing a loss function.

## Tensor Data Structures

1. To represent very general types of data, we rely on $D$-dimensional tensors
(multidimensional arrays).  Vectors and matrices are special 1 and 2D cases of
tensors. Classes $y_i \{1, \dots, K\}$ can be viewed as one-hot encoded vectors.
Therefore, deep learning immediately applies to the data considered in more
traditional statistics problems.

1. Images and video can be viwed as 3 and 4D tensors, respectively. An RGB image
has dimensions Height $\times$ Width $\times$ Color Channel. A video adds an
extra Time dimension.

1. A document can be viewed as a one-hot encoded tensor with dimensions Vocabulary Size $\times$ Document Length. Notice that since documents can have different lengths, a typical dataset will include many tensors with different sizes.

1. We can stack tensors into "batches" of data points $x_i$. Often, these will
themselves form tensors. For example, a batch of 10 RGB images, each of which is
16 pixels tall and wide, can be written as a $10 \times 16 \times 16 \times 3$
tensor. This property is helpful for accelerating computation. Since GPUs can
perform tensor arithmetic in parallel, we'll be able to optimize over entire
batches very efficiently.

## Perceptrons

1. We'll eventually want flexible modules for mapping tensors from input to
output. Optimizing over those modules will allow us to achieve our prediction
objective.  But for now, let's consider how to map a vecor $x \in \reals^{D}$ to
a binary label $y \in \{0, 1\}$ using the classic perceptron algorithm.

1.

## Multilayer Networks

## Parameters and Activations

## Types of Layers

1. **Linear Layers**.

1. **Activation Layers**

1. **Normalization Layers**

1. **Output Layers**

## Learning Representations

## Code Example