---
title: "Partial Dependence Profiles II"
description: |
author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: 2021-04-12
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library("knitr")
opts_chunk$set(cache = TRUE, message = FALSE, warning = FALSE, echo = TRUE)
```

_[Reading](), [Recording](), [Rmarkdown]()_

```{r}
library("ggplot2")
library("DALEX")
theme479 <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    legend.position = "bottom"
  )
theme_set(theme479)
```

1. Partial dependence (PD) plots help answer the question, « How is feature $j$
used by my model $f$? » Slight variations on PD plots are useful for some
related followup questions,
  	* Has my model $f$ learned interactions between features $j$ and $j^{\prime}$?
  	* How do the models $f$ and $f^{\prime}$ differ in the way that they use feature $j$?
  	
2. The variants, called Grouped and Contrastive PD plots, reduce the original CP
profiles less aggressively than PD plots, but without becoming overwhelmingly
complicated.

### Interactions

3. We say that there is an interaction between variables $j$ and $j^{\prime}$
if the relationship between $x\_{j}$ and $y$ is modulated by variable
$j^{\prime}$. For example, in the figure below, the slope of cross-sections
across $j$ depends on $j^{\prime}$.

4. Using the language of CP profiles, the figure above means that the shape of
the CP profile in $j$ depends on the particular setting of $j^{\prime}$. This
motivates the use of Grouped PD profiles — we compute several PD profiles in
$j$, restricting attention to CP profiles whose value $x\_{j^{\prime}}$ lies
within a prespecified range.

5. For example, consider the housing dataset. We’ve faceted the CP profiles in
$j$ depending on ….

```{r, fig.cap = "Grouping the CP profiles for average occupancy by home age reveals an interaction effect."}
fit <- get(load("~/Desktop/fit.rda"))
data(titanic)
titanic <- titanic %>%
  select(-country) %>%
  na.omit()
x <- select(titanic, -survived)

explanation <- explain(model = fit, data = x, y = titanic$survived)
profiles <- model_profile(explainer = explanation, groups = "class")
plot(profiles, geom = "profiles", variables = "age") +
  scale_color_brewer(palette = "Set2") +
  theme479
```

6. What should we do if there are many input variables and we don’t have a
priori knowledge about which variables $j^{\prime}$ might be interacting with
$j$? One idea is to try to *discover* relevant interactions by clustering the
original set of CP profiles.

```{r}
profiles <- model_profile(explainer = explanation, variables = "age", k = 3)
plot(profiles, geom = "profiles", variables = "age") +
  scale_color_brewer(palette = "Set2") +
  theme479
```

7. The idea is to compute the CP profiles for all the samples, and then see
whether there are subsets of profiles that all look similar. If we can find
features $j^{\prime}$ that characterize these groupings, then we have found
features that interact with $j$ (with respect to the fitted model $f$).

### Model Comparison

8. The comparison of different models’ PD profiles can be used to,
  	* Validate a simple model
  	* Guide the design of new features, and
  	* Characterizing overfitting
  	PD profiles that are used to compare different models are sometimes called
  	« Contrastive » PD profiles.
  	
9. To validate a simple model, we can compare its PD profiles with those of a
more sophisticated model. For example, in the display below, the random forest
learns linear relationships with price for both the surface and floor variables.
If all the effects were like this, then we would have a good reason for
preferring the linear model.

```{r}
data(apartments)
x <- select(apartments, -m2.price)
profiles_lm <- train(x, apartments$m2.price, method = "lm") %>%
  explain(x, apartments$m2.price, label = "LM") %>%
  model_profile()
profiles_rf <- train(x, apartments$m2.price, method = "rf", tuneGrid = data.frame(mtry = 10)) %>%
  explain(x, apartments$m2.price, label = "RF") %>%
  model_profile()
```

```{r}
plot(profiles_lm, profiles_rf, variables = c("surface", "floor")) +
  scale_color_brewer(palette = "Set2") +
  theme479
```

10. When making the comparison between a simple and a complex model, certain
discrepancies might become apparent. For example, important nonlinearities or
interactions might be visible from the PD profiles of the complex model. This
information can guide the design of new features in the simpler model, so that
it can continue to be used. This is exactly the case in the apartments dataset
above -- there is a strong nonlinear relationship for the construction year
variables. This suggests that, if a linear model is still desired, then a new
feature should be defined that identifies whether the apartment was built
between 1935 and 1990.

```{r}
plot(profiles_lm, profiles_rf, variables = "construction.year") +
  scale_color_brewer(palette = "Set2") +
  theme479
```

11. Suppose you have found that a model is overfitting (e.g., by finding that
it’s training error is much lower than its test error). One way to address this
overfitting is to compare the PD profiles between the simple and complex models.
If the profiles are very different for one of the features, then that feature
may be the source of overfitting.
