---
title: "Raster Data"
description: |
  Storing spatially gridded information in rasters.
author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: 03-10-2021
output:
  distill::distill_article:
    self_contained: false
---

_[Reading](https://geocompr.robinlovelace.net/spatial-class.html), [Recording](https://mediaspace.wisc.edu/media/Week%207%20%5B3%5D%20Raster%20Data/1_k5px5f53), [Rmarkdown](https://github.com/krisrs1128/stat479/blob/master/_posts/2021-03-02-week7-3/week7-3.Rmd)_

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

```{r}
library("raster")
library("RStoolbox")
library("ggplot2")
library("dplyr")
theme_set(theme_bw())
```

1. The raster data format is used to store spatial data that lie along regular
grids. The values along the grid are stored as entries in the matrix. The raster
object contains metadata that associates each entry in the matrix with a
geographic coordinate.

2. Since the data they must lie along a regular grid, rasters are most often
used for continuously measured data, like elevation, temperature, population
density, or landcover class.

3. We can create a single layer raster using the `raster` command. The code
block below loads an elevation map measured by the space shuttle.

```{r}
f <- system.file("raster/srtm.tif", package = "spDataLarge")
zion <- raster(f)
```

4. Typing the name of the object shows the metadata associated with it (but not
the actual grid values). We can see that the grid has 457 rows and 465 columns.
We also see its spatial extent: The minimum and maximum longitude are both close
to -113 and the latitudes are between 37.1 and 37.5. A quick google map [search](https://goo.gl/maps/X4EY2BsGgMZzN5BM9) shows that this is located in Zion
national park.

```{r}
zion
```
```{r}
plot(zion)
```

5. The `raster` command also lets us create raster objects from scratch. For example, the code below makes a raster with increasing values in a 6 x 6 grid. Notice that we had to give a fake spatial extent.

```{r}
test <- raster(
  nrows = 6, ncols = 6, res = 0.5, 
  xmn = -1.5, xmx = 1.5, ymn = -1.5, ymx = 1.5,
  vals = 1:36
)

plot(test)
```

6. Real-world rasters typically have more than one layer of data. For example, you might measure both elevation and slope along the same spatial grid, which would lead to a 2 layer raster. Or, for satellite images, you might measure light at multiple wavelengths (usual RGB, plus infrared or thermal for example).

7. Multi-layer raster data can be read in using `brick`. You can refer to particular layers in a multi-layer raster by using `subset`.

```{r}
f <- system.file("raster/landsat.tif", package = "spDataLarge")
satellite <- brick(f)
satellite
```

8. Base R’s `plot` function supports plotting one layer of a raster at a time. To plot more than one layer in a multichannel image (like ordinary RGB images) you can use the `plotRGB` function.

```{r}
plotRGB(satellite)
```

9. Sometimes, it’s useful to overlay several visual marks on top of a raster
image. In this case, the `ggRGB` function from the `RStoolbox` package can be
used.

```{r}
satellite <- projectRaster(satellite, crs = 4326)
ggRGB(satellite) +
  geom_point(
    data = data.frame(x = -113, y = 37.3), 
    aes(x = x, y = y), 
    col = "red", size = 10
  )
```

10. If we want to visualize just a single layer using ggplot2, we can use
`geom_raster`. This assumes that the data are in "tall" format, which can be
accomplished by (1) subsetting to the relevant layer and (2) converting it to a
data.frame. We use `xy = TRUE` to remember the original locations of each pixel.

```{r}
sat_df <- subset(satellite, 1) %>%
  as.data.frame(xy = TRUE)
ggplot(sat_df) +
  geom_raster(aes(x = x, y = y, fill = landsat.1)) +
  scale_fill_gradient(low = "white", high = "black") +
  coord_fixed()
```

