---
title: "Introduction to Feature Learning"
description: | 
author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: 2021-04-19
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library("knitr")
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_[Reading](), [Recording](), [Rmarkdown]()_

```{r}
library("keras")
```

1. In classical machine learning, we assume that the features most relevant to
prediction are already available. E.g., when we want to predict home price, we
already have features about square feet and neighborhood income, which are
clearly relevant to the prediction task.

2. In many modern problems though, we have only access to « raw » data where the
most relevant features are not directly encoded.

  	* In image classification, we only have raw pixel values. Say we want to
  	identify where the pedestrians are in an image.
  	* For sentiment analysis, we want to identify whether a piece of text is a
  	positive or negative valence. We only have access to the raw sequence of words.
	
3. The features that are truly relevant only arise as complex interactions
between these elements.
  
  	* To recognize a pedestrian, need many adjacent pixels to have a certain
  	configuration of values, each which becomes recognizable as a part of a
  	person.
  	* To recognize a positive or negative sentiment, need to recognize
  	interactions between words. "The movie was good until" clearly has bad
  	sentiment, but you cannot tell that from the isolated word counts alone.
	
4. The main idea of deep learning is to learn these more complex features one
layer at a time.

	1. For image data, it has this structure.
	2. The first layer recognizes interactions between individual pixels.
	Specifically, individual features are designed to « activate » when particular
	pixel interactions are present.
	3. The second layer learns to recognize interactions between features in the
	first layer.
	4. And so on, until the learned features correspond to more « high-level »
	concepts, like sidewalk or pedestrian.
	5. Notion of feature maps. Large value when it recognizes a feature within a
	given spatial context.
	
5. At the end of the feature extraction process, all the features are passed
into a final linear or logistic regression module that completes the regression
or classification task, respectively.

6. It is common to refer to each feature map as a « neuron. » Different neurons
activate when different patterns are present in the original, underlying image.

7. Illustrate this on real dataset
	1. Load a model and show the shapes of the features.
	2. There are many types of layers shown here. Each type of layer contributes in
	a different way to the feature learning goal, and learning how design and
	compose these different types of layers is one of the central concerns of deep
	learning (we won’t touch on this any further).
	3. We will see how to load and inspect the features in the next lecture

```{r}
f <- tempfile()
download.file("https://uwmadison.box.com/shared/static/9wu6amgizhgnnefwrnyqzkf8glb6ktny.h5", f)
model <- load_model_hdf5(f)
model
```
8. This idea is not limited to image data.
	1. In sentiment analysis, first layer features activate when specific
	interactions between words are present
	2. The next layer learns interactions between phrases,…
	3. In the handwriting example, learn the interactions between points where the
	pen was registered.
	4. Then new features that activate when particular curves or handwriting
	strokes are present.
	
9. Deep learning is often called a black box because these intermediate features
are often complex and not directly interpretable according to human concepts
	1. This is further complicated by the fact that features are « distributed »
	2. Individual human concepts are not associated with individual deep learning
	features. Instead, a single human concept is encoded by a configuration of
	multiple features. On the flipside, the same model feature can encode several
	human concepts.
	
10. Nonetheless, a growing literature addresses the question of model
interpretability by attempting to visualize and interact with these learned
features.