---
title: "Pointwise Diagnostics"
description: |
author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: 2021-04-12
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library("knitr")
opts_chunk$set(cache = TRUE, message = FALSE, warning = FALSE, echo = TRUE)
```

_[Reading](), [Recording](), [Rmarkdown]()_

```{r}
library("dplyr")
library("ggplot2")
library("ggrepel")
library("loo")
library("purrr")
library("rstan")
library("tidyr")
theme479 <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    legend.position = "bottom"
  )
```

1. All the model visualization strategies we’ve looked at in the last few
lectures have been dataset-wide. That is., we looked at properties of the
dataset as a whole, and whether the model made sense globally, across the whole
dataset. Individual observations might warrant special attention, though.

2. A first diagnostic to consider is the leave-one-out predictive distribution.
This is the probability $p\left(y_{i} \vert y_{-i}\right)$ of sample $i$ after
having fitted a model to all samples except $i$. Ideally, most observations in
the dataset to have high predictive probability.
    * Note that this can be used for model comparison. Some models might have better
    per-sample leave-one-out predictives than others.
		*  This is similar to a leave-one-out residual.
		
```{r}
models <- get(load("~/Desktop/models.rda"))
elpd <- map(models, ~ loo(., save_psis = TRUE)$pointwise[, "elpd_loo"])
elpd_diffs <- GM@data %>%
  mutate(
    ID = row_number(),
    diff23 = elpd[[3]] - elpd[[2]]
  )

outliers <- elpd_diffs %>%
  filter(abs(diff23) > 6)
```

```{r}
ggplot(elpd_diffs, aes(ID, diff23)) +
  geom_point(
    aes(col = super_region_name),
    size = 0.9, alpha = 0.8
    ) +
  geom_text_repel(
    data = outliers,
    aes(label = City_locality),
    size = 3 
  ) +
  labs(
    y = "Influence (Model 2 vs. 3)",
    col = "WHO Region"
  )
```

3. Another diagnostic is to consider the influence of an observation. Formally,
the influence is a measure of how much the posterior predictive distribution
changes when we leave one sample out. The idea is to measure the difference
between the posterior predictives using a form of KL divergence, and note down
the observations that lead to a very large difference in divergence.

```{r}
loglik <- map(models, ~ as.matrix(., pars = "log_lik"))
k <- psis(loglik[[2]])$diagnostics$pareto_k
kdata <- data.frame(Index = seq_along(k), k_hat = k)
outliers <- kdata %>%
  filter(k_hat > 0.25)
ggplot(kdata, aes(x = Index, y = k_hat)) + 
  geom_point(size = 0.5, alpha = 0.9) + 
  geom_text_repel(data = outliers, aes(label = Index)) +
  labs(y = "k-hat")
```
